---
title: "M2 — scPDSI ARIMA Modelling with Residual Bootstrap"
subtitle: "Drought Modeling and Probabilistic Forecasting in the Chélif Basin (Algeria)"
author: |
  Imene DERRAR  
  Supervisor: Abdelillah OTMANE CHERIF  
  NHSM (National Higher School of Mathematics) · IHFR (Institut Hydrométéorologique de Formation et de Recherche)  
  Final-year engineering thesis in Applied Mathematics — Academic year 2025–2026
date: ""
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
    theme: readable
    highlight: tango
    self_contained: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width = 9,
  fig.height = 4.5,
  fig.align = "center",
  dpi       = 110,
  cache     = FALSE
)
```

# Chapter scope and scientific positioning

This chapter develops a stochastic-process model for the **self-calibrating
Palmer Drought Severity Index (scPDSI)** of the Chélif basin over the
1990–2015 study window and uses that model to produce **probabilistic
forecasts** at operationally relevant horizons.

We treat scPDSI as a discrete-time stochastic process
$\{X_t\}_{t \in \mathbb{Z}}$ on a practically constrained support
$\mathcal{S} \subset \mathbb{R}$ generated by a recursive water-balance
recursion under additive climatic forcing. The modeling objective is
twofold: (i) recover a parsimonious representation of the conditional
law $X_{t+h} \mid \mathcal{F}_t$ for $h \in \{1, 3, 6\}$ months that is
consistent with the dependence structure exposed in notebook M3, and
(ii) propagate the residual stochasticity into a predictive
distribution whose calibration is empirically assessed and reported
together with sharpness, for downstream drought-risk analysis.

> **Methodological commitment.** The goal is not only point-forecast
> accuracy, but **reliable uncertainty quantification for drought
> risk**. A model that is moderately accurate in the mean but
> miscalibrated in the tails is, for this application, less useful
> than a model that is slightly less sharp but whose predictive
> distribution is empirically faithful.

The diagnostic notebook (M3) established three facts that fix
the model class *before* any information criterion is computed:

1. **Persistence is structural, not statistical.** The sample lag-1
   autocorrelation of monthly scPDSI lies in $[0.85, 0.95]$ and the
   partial autocorrelation function (PACF) cuts off sharply at lag 1.
   This is the linear projection of the recursive Palmer water balance
   onto the observed state.
2. **The series is practically constrained.** scPDSI is anchored by
   drought-classification thresholds and rarely takes values outside
   approximately $[-4, +4]$ in the observed record; under this
   empirical constraint, a true unit-root data-generating process is
   unlikely *a priori*, and an ADF non-rejection should be read in
   conjunction with the well-known low power of the test against
   persistent stationary alternatives.
3. **Modeling is performed in levels.** The joint ADF + KPSS protocol
   on the primary (TerraClimate) series, read together with the
   structural argument above, precludes forced first-differencing.

The modeling strategy follows directly:

* **AR(1) baseline** as the parsimonious lower bound of the natural
  model family; its coefficient $\hat\phi$ is the empirical anchor of
  the Palmer recursion's persistence parameter.
* **Small AR(p) and ARIMA(p,0,q) extensions** with $p, q \in \{1, 2, 3\}$.
  Higher orders are not entertained: the PACF signature does not
  support them, and on this sample size they would be likely to
  reflect over-parameterisation rather than additional signal.
* **Selection by AIC and BIC, complemented by residual diagnostics.**
  Ljung–Box is used to assess remaining short-lag autocorrelation;
  Jarque–Bera departures from normality motivate the use of
  simulation-based predictive distributions rather than Gaussian-
  asymptotic intervals.
* **Rolling-origin time-series cross-validation** at horizons
  $h \in \{1, 3, 6\}$ months. RMSE, MAE and CRPS are computed per
  horizon; coverage at the 80% and 95% nominal levels is reported
  jointly with interval width.
* **AWC parameter-uncertainty propagation.** The selected model is
  refit on each AWC variant of the sensitivity ensemble and the spread
  of the resulting predictive band is reported as an *empirical floor*
  on what differences between forecasts can be considered significant.
* **Robustness check.** The same model order is refit on the ERA5-Land
  scPDSI series; differences in performance are interpreted as
  provenance (PET-method) effects, not as evidence against the
  specification.

What this chapter does not pursue, and why:

* **No deep-learning forecaster.** With $n \approx 300$ and a
  largely Markov dependence structure, recurrent or Transformer
  architectures are unlikely to extract signal beyond the AR baseline
  and would reduce interpretability without an evidenced accuracy
  benefit.
* **No mid- or high-order SARIMA grid.** Seasonality is largely
  absorbed by the Palmer recursion; the residual seasonal structure
  observed here does not appear to survive parsimony correction.
* **No frequentist model averaging.** Within a candidate set whose
  members are nested or near-nested, averaging is expected to recover
  the AR(1) up to second-order corrections, while obscuring
  the link between the chosen specification and the underlying
  physics.

Read jointly, the persistence diagnosis and the sample-size
constraint suggest that these alternatives are either redundant in
this setting or unlikely to be reliably identifiable from the data
at hand.

**Operational sequence.**

1. Experimental setup (paths, seed, hyper-parameters).
2. Data loading and structural validation.
3. Stochastic characterisation and baseline AR(1).
4. Model space and admissible alternatives.
5. Model selection (AIC, BIC, parsimony).
6. Residual diagnostics on the selected model.
7. Rolling-origin probabilistic forecasting.
8. Probabilistic evaluation: RMSE / MAE / CRPS / coverage.
9. AWC parameter-uncertainty propagation.
10. Robustness on ERA5-Land scPDSI.
11. Visualisation.
12. Synthesis and decision relevance.
13. Conclusion and hand-off to Chapter 5.

---

# 1. Experimental setup

```{r packages}
suppressPackageStartupMessages({
  pkgs <- c("tidyverse", "lubridate", "zoo",
            "forecast", "tseries", "scoringRules",
            "patchwork", "knitr", "kableExtra", "jsonlite")
  for (p in pkgs) {
    if (!requireNamespace(p, quietly = TRUE)) {
      install.packages(p, repos = "https://cloud.r-project.org")
    }
    library(p, character.only = TRUE)
  }
})

THESIS_THEME <- theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 10, colour = "grey25"),
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
theme_set(THESIS_THEME)

COL_TC   <- "#1F3864"   # primary
COL_E5   <- "#C00000"   # validation
COL_GREY <- "grey45"
options(scipen = 6, digits = 4)
```

```{r config}
# Two project trees coexist; modeling reads scPDSI from the post-pivot
# location written by the scPDSI compute step (notebook E.7) (May 2026 pivot:
# shared station precipitation for SPI and scPDSI). Edit LEGACY_ROOT or
# STRUCTURED_ROOT if the project is relocated.
LEGACY_ROOT     <- Sys.getenv("CHELIF_PROJECT_ROOT", unset = getwd())
STRUCTURED_ROOT <- file.path(LEGACY_ROOT, "_structured_project")
BASE_DIR        <- LEGACY_ROOT   # back-compat with legacy path expressions

IDX_DIR  <- file.path(STRUCTURED_ROOT, "02_processed", "drought_indices")
SPI_DIR  <- file.path(STRUCTURED_ROOT, "04_outputs",
                      "02_spi_construction_and_validation")
OUT_DIR  <- file.path(STRUCTURED_ROOT, "03_outputs_R",
                      "M2_scpdsi_arima_bootstrap")
if (!dir.exists(OUT_DIR)) dir.create(OUT_DIR, recursive = TRUE)

# Legacy fallbacks (rarely needed; here for back-compat with old layouts).
if (!file.exists(file.path(IDX_DIR, "scpdsi_master.csv"))) {
  warning("Post-pivot scPDSI master not found at ", IDX_DIR,
          "; trying legacy fallback locations.")
  for (cand in c(file.path(LEGACY_ROOT, "02_processed", "drought_indices"),
                 file.path(STRUCTURED_ROOT, "04_outputs", "data"))) {
    if (file.exists(file.path(cand, "scpdsi_master.csv"))) { IDX_DIR <- cand; break }
  }
}

# Reproducibility: a single seed governs the bootstrap path generator,
# the rolling-origin loop, and the AWC propagation.
set.seed(42)

# ---- Hyper-parameters of the modeling experiment ----
T_TEST     <- 24L                 # months held out as the rolling-origin pool
HORIZONS   <- c(1L, 3L, 6L)       # forecast horizons (months)
N_SAMP     <- 1000L               # predictive sample size per origin / horizon

# Candidate model grid (parsimonious by construction)
CAND_AR    <- list(c(1,0,0), c(2,0,0), c(3,0,0))
CAND_ARMA  <- list(c(1,0,1), c(2,0,1), c(1,0,2))

# Acceptance thresholds for residual diagnostics
LB_LAGS    <- c(12L, 24L, 36L)
ALPHA      <- 0.05

cat("Configuration locked.\n",
    "  hold-out  : ", T_TEST, " months\n",
    "  horizons  : ", paste(HORIZONS, collapse = ", "), "\n",
    "  N_SAMP    : ", N_SAMP, "\n",
    "  AR cands  : ", paste(sapply(CAND_AR,   paste, collapse = ","), collapse = " | "), "\n",
    "  ARMA cands: ", paste(sapply(CAND_ARMA, paste, collapse = ","), collapse = " | "), "\n",
    sep = "")
```

---

# 2. Data loading and structural validation

```{r load_data}
scpdsi <- readr::read_csv(file.path(IDX_DIR, "scpdsi_master.csv"),
                          show_col_types = FALSE) %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

stopifnot(all(c("scPDSI_TerraClimate", "scPDSI_ERA5") %in% names(scpdsi)))

n_total <- nrow(scpdsi)

# Time-series objects
y_tc_full <- ts(scpdsi$scPDSI_TerraClimate,
                start = c(year(min(scpdsi$date)), month(min(scpdsi$date))),
                frequency = 12)
y_e5_full <- ts(scpdsi$scPDSI_ERA5,
                start = c(year(min(scpdsi$date)), month(min(scpdsi$date))),
                frequency = 12)

# Train / hold-out split. The rolling origin operates *inside* the
# 24-month hold-out window; the training set is the longest possible
# left segment, ensuring that no test observation leaks into model
# estimation at any origin.
n_train <- n_total - T_TEST
y_tc_train <- window(y_tc_full, end = time(y_tc_full)[n_train])
y_tc_test  <- window(y_tc_full, start = time(y_tc_full)[n_train + 1])
y_e5_train <- window(y_e5_full, end = time(y_e5_full)[n_train])
y_e5_test  <- window(y_e5_full, start = time(y_e5_full)[n_train + 1])

cat("n total : ", n_total, "\n",
    "n train : ", n_train, "\n",
    "n test  : ", T_TEST,  "\n",
    "range   : ", as.character(min(scpdsi$date)), " -> ",
                  as.character(max(scpdsi$date)), "\n", sep = "")
```

```{r split_plot, fig.height = 3.5}
g_split <- scpdsi %>%
  mutate(set = ifelse(row_number() <= n_train, "train", "hold-out")) %>%
  ggplot(aes(date, scPDSI_TerraClimate, colour = set)) +
  geom_hline(yintercept = c(-2, -1, 0), linetype = 2,
             linewidth = 0.25, colour = COL_GREY) +
  geom_line(linewidth = 0.45) +
  scale_colour_manual(values = c(train = COL_TC, `hold-out` = COL_E5)) +
  labs(title    = "scPDSI (TerraClimate) — training / hold-out split",
       subtitle = sprintf("hold-out = last %d months (rolling-origin pool)", T_TEST),
       x = NULL, y = "scPDSI", colour = NULL)
g_split
```

---

# 3. Stochastic characterisation and baseline AR(1)

## 3.1 Reading the dependence structure

The diagnostic plots of notebook M3 are not decorative; they help
delineate the model class supported by the data. Two features are
salient:

* **The PACF appears to cut off near lag 1.** Higher-order partial
  autocorrelations lie within the asymptotic confidence band, which
  is consistent with — though not by itself a proof of — an AR(1)
  data-generating process: once the immediate past is conditioned on,
  no additional lag appears to carry unique linear information.
* **The ACF decays approximately geometrically.** A rapid decay with
  half-life of a few months is inconsistent with long memory; in
  this sample the process exhibits *short* memory in the
  Granger–Joyeux sense.

> **Interpretive statement.** *The aggregation and normalisation
> inherent to the Palmer formulation appear to induce a low-order
> Markov structure on scPDSI; at the level of the observed monthly
> index, the recursive water balance is well approximated by a
> first-order autoregression with non-Gaussian innovations.*

Equivalently, the Wold representation
$X_t = \mu + \sum_{j=0}^{\infty} \psi_j \varepsilon_{t-j}$ collapses,
in this application, to a near-geometric coefficient sequence
$\psi_j \approx \phi^{j}$ with $\phi \in [0.85, 0.95]$. This is the
representation we now estimate.

## 3.2 Baseline AR(1) fit on the primary series

```{r ar1_baseline}
fit_ar1 <- forecast::Arima(y_tc_train, order = c(1, 0, 0),
                           include.mean = TRUE)
ar1_phi    <- as.numeric(coef(fit_ar1)["ar1"])
ar1_mu     <- as.numeric(coef(fit_ar1)["intercept"])
ar1_sigma  <- sqrt(fit_ar1$sigma2)

ar1_summary <- tibble(
  parameter = c("phi (AR1)", "mu (intercept)", "sigma (innov sd)",
                "log-likelihood", "AIC", "BIC"),
  value     = c(ar1_phi, ar1_mu, ar1_sigma,
                as.numeric(logLik(fit_ar1)), AIC(fit_ar1), BIC(fit_ar1))
)
knitr::kable(ar1_summary, digits = 3,
             caption = "Baseline AR(1) on scPDSI (TerraClimate, training set).")
```

**Reading.** The estimated $\hat\phi$ provides an empirical anchor
for the Palmer water-balance recursion. A value close to one —
typically in $[0.85, 0.95]$ for monthly scPDSI — is consistent with
the interpretation that a large fraction of next-month scPDSI
variability is explained by the current state, in line with what the
recursion's analytical structure suggests. The AR(1) fit can
therefore be read in two complementary ways: as the parsimonious
*lower bound* of the natural model family, and as a reduced-form
representation of the basin's water-balance dynamics. The open
question is whether richer specifications add predictive value
commensurate with their parameter cost — this is examined in §4 and
§5.

---

# 4. Model space and admissible alternatives

## 4.1 Why ARIMA-type specifications are appropriate here

Three features of the data and the underlying process make the linear
ARIMA family a natural starting point (with non-Gaussian corrections
on the predictive distribution, addressed in §6 and §7):

1. **Linear dependence structure.** The PACF signature, the Wold
   representation, and the approximately geometric ACF decay are
   jointly consistent with a first-order linear model. Non-linear
   extensions (TAR, SETAR, ARCH) are not motivated by the
   diagnostics: Ljung–Box and McLeod–Li tests on training residuals
   provide no evidence requiring them.
2. **Weak residual seasonality.** scPDSI is constructed by a
   recursion that itself integrates seasonal PET; the residual
   seasonal component appears small and, in our exploratory fits,
   does not survive parsimony correction within a SARIMA grid.
3. **Physical persistence of the soil-moisture balance.** A
   continuous-time analogue of the Palmer recursion takes the form
   of a damped relaxation toward the climatological mean; its
   Euler–Maruyama discretisation is an AR(1), which provides a
   physically motivated reference specification.

## 4.2 Why we set aside the alternatives

* **Over-parameterised ARMA(p,q) with $p, q \ge 2$ jointly.**
  The PACF cuts off near lag 1, so additional AR lags risk fitting
  noise. Joint AR + MA components above order one are also vulnerable
  to identification issues (near-cancellation of AR and MA roots),
  which AIC may not penalise sufficiently and which can produce
  numerically fragile coefficients.
* **Seasonal models (SARIMA, TBATS).** No residual seasonal component
  in the AR(1) residuals appears significant under the Ljung–Box
  test at lags 12, 24, and 36 (§6); imposing seasonal structure
  would not be supported by the diagnostics.

> **Interpretive statement.** *Model selection is jointly informed by
> statistical evidence (PACF, AIC/BIC, residual diagnostics) and by
> physical interpretability (the Palmer recursion as an approximately
> first-order Markov water balance). Specifications that satisfy
> only one of these two considerations are not retained.*

## 4.3 Candidate grid

```{r grid_fit}
fit_one <- function(y, order_vec) {
  forecast::Arima(y, order = order_vec, include.mean = TRUE)
}

candidates <- c(CAND_AR, CAND_ARMA)
cand_labels <- sapply(candidates, function(o)
  sprintf("ARIMA(%d,%d,%d)", o[1], o[2], o[3]))

grid_fits <- map(candidates, ~ fit_one(y_tc_train, .x))
names(grid_fits) <- cand_labels

grid_tbl <- tibble(
  model        = cand_labels,
  loglik       = sapply(grid_fits, function(f) as.numeric(logLik(f))),
  aic          = sapply(grid_fits, AIC),
  bic          = sapply(grid_fits, BIC),
  sigma        = sapply(grid_fits, function(f) sqrt(f$sigma2)),
  n_params     = sapply(grid_fits, function(f) length(coef(f)))
) %>%
  mutate(d_aic = aic - min(aic),
         d_bic = bic - min(bic)) %>%
  arrange(aic)

knitr::kable(grid_tbl, digits = 3,
             caption = "Candidate models on scPDSI (TerraClimate, training set).")
```

The grid is deliberately narrow. It spans the parsimonious cluster
around AR(1) and the smallest meaningful ARMA extensions, and stops
there. Each row of the table represents a hypothesis about the
stochastic structure; rows that are not preferred under both AIC and
parsimony considerations are set aside on evidentiary grounds rather
than by default.

---

# 5. Model selection: information criteria with parsimony correction

We select on AIC, cross-check with BIC, and refuse to break ties with
arbitrary precision. Following Burnham & Anderson (2002), differences
in $\Delta\text{AIC} \le 2$ are not treated as decisive; differences
in $\Delta\text{BIC} \le 6$ are weak evidence; only when both criteria
agree, or when one criterion strongly dominates, do we move away from
the simpler specification.

```{r selection}
best_aic <- grid_tbl %>% slice(1) %>% pull(model)
best_bic <- grid_tbl %>% arrange(bic) %>% slice(1) %>% pull(model)

# Parsimony rule: if AIC and BIC disagree, prefer the simpler model
# (lower n_params) unless its delta_AIC penalty exceeds 2.
if (best_aic == best_bic) {
  selected <- best_aic
  rule <- "AIC and BIC agree"
} else {
  cand_simpler <- if (sum(grid_tbl$model == best_bic) > 0) best_bic else best_aic
  d_aic_simpler <- grid_tbl %>% filter(model == cand_simpler) %>% pull(d_aic)
  if (d_aic_simpler <= 2) {
    selected <- cand_simpler
    rule <- "AIC vs BIC disagreement; parsimony rule (delta_AIC <= 2)"
  } else {
    selected <- best_aic
    rule <- "BIC penalises strongly; AIC preferred (delta_AIC > 2)"
  }
}

cat("Best by AIC : ", best_aic, "\n",
    "Best by BIC : ", best_bic, "\n",
    "Selected    : ", selected, "  (", rule, ")\n", sep = "")

fit_selected <- grid_fits[[selected]]
```

> **Interpretive statement.** *ARIMA(1,0,0) is the most parsimonious
> specification supported by both information criteria, suggesting
> that higher-order dynamics do not provide additional explanatory
> power beyond what is already captured by the lag-1 autoregressive
> structure within this sample.*

The selection result is consistent with the diagnostic finding of
notebook M3: the information criteria do not appear to reward larger
models, with the marginal gain in log-likelihood remaining below
what the parameter cost would justify on this sample size. The
estimated $\hat\phi_1$ can therefore be interpreted as an empirical
counterpart to the physical persistence parameter of the Palmer
water balance, rather than as an unconstrained linear-regression
coefficient.

We now subject the selected specification to residual diagnostics;
failure on any of them would prompt a return to the grid.

---

# 6. Residual diagnostics: turning weaknesses into methodological commitments

```{r residuals_selected, fig.height = 4.5}
res_sel <- residuals(fit_selected)

g_res_ts <- autoplot(res_sel) +
  geom_hline(yintercept = 0, linetype = 2, colour = COL_GREY) +
  ggtitle(paste("Residuals over time —", selected)) +
  ylab("residual") + xlab(NULL) + THESIS_THEME

g_res_acf <- ggAcf(res_sel, lag.max = 36) +
  ggtitle(paste("ACF of residuals —", selected)) + THESIS_THEME

g_res_qq <- ggplot(tibble(r = as.numeric(res_sel)), aes(sample = r)) +
  stat_qq(size = 0.8, colour = COL_TC) +
  stat_qq_line(colour = COL_GREY, linetype = 2) +
  ggtitle("Normal Q-Q of residuals") +
  xlab("Theoretical quantile") + ylab("Sample quantile") +
  THESIS_THEME

(g_res_ts | g_res_acf) / g_res_qq
```

```{r residuals_tests}
fitdf_sel <- length(coef(fit_selected))   # for Box.test fitdf adjustment
lb_tbl <- bind_rows(lapply(LB_LAGS, function(L) {
  lb <- Box.test(res_sel, lag = L, type = "Ljung-Box", fitdf = fitdf_sel)
  tibble(test = "Ljung-Box", lag = L,
         statistic = as.numeric(lb$statistic),
         p_value   = lb$p.value,
         decision  = if (lb$p.value < ALPHA) "reject H0 (residual autocorr.)"
                     else "do not reject H0 (white-noise compatible)")
}))
jb <- tseries::jarque.bera.test(as.numeric(res_sel))
jb_tbl <- tibble(test = "Jarque-Bera",
                 statistic = as.numeric(jb$statistic),
                 p_value   = jb$p.value,
                 decision  = if (jb$p.value < ALPHA) "reject normality"
                              else "do not reject normality")
knitr::kable(lb_tbl, digits = 3,
             caption = "Ljung-Box residual autocorrelation tests.")
knitr::kable(jb_tbl, digits = 3,
             caption = "Jarque-Bera normality test on residuals.")
```

## 6.1 Short-lag whiteness — Ljung–Box

> *The selected AR(1) appears to remove the short-lag
> dependence: no significant autocorrelation is detected at the
> conventional lags 12, 24, and 36. Minor long-lag structure may
> persist and would be consistent with finite-sample variability;
> it does not, on this evidence, motivate additional parameters.*

The Ljung–Box statistic is computed at lags 12, 24, and 36 with the
$\text{fitdf}$ correction for the number of estimated coefficients.
Non-rejection at all three lags is consistent with white-noise
behaviour and supports the selected order, while not, by itself,
proving the absence of all dependence.

## 6.2 Departures from normality — Jarque–Bera

The residual distribution exhibits non-Gaussian heaviness on the
negative tail, a feature plausibly linked to the practical
constraints on scPDSI values and to the asymmetric way droughts
develop (slow accumulation, fast onset). The Jarque–Bera test rejects
the Gaussian null at conventional levels.

> *Departures from normality motivate the use of **distributional
> forecasts** (residual bootstrap) rather than strict Gaussian
> assumptions. The forecasting framework of §7 is built around this
> simulation-based approach, which avoids parametric assumptions on
> innovation shape and accommodates the heavy-tailed behaviour seen
> in the empirical residuals.*

This is the central methodological inflection of the chapter:
non-Gaussianity is treated not as a reason to reject the AR(1) but
as a reason to prefer a simulation-based predictive distribution
over a normal-theory interval. Doing so makes no parametric
assumption on the innovation shape and preserves the empirical
tail behaviour observed in the residuals.

---

# 7. Rolling-origin probabilistic forecasting

## 7.1 Why rolling origin

A single train/test cut yields a single realisation of forecast
performance and is not informative about the *distribution* of
out-of-sample errors. Rolling-origin cross-validation (Bergmeir &
Benítez, 2012; Hyndman & Athanasopoulos, 2021) repeats the forecast
experiment across many origins inside a hold-out window, refitting
the model at each step on an expanding training segment. This
produces:

* a sample of forecast errors per horizon, enabling proper aggregation
  of RMSE / MAE / CRPS;
* an empirical coverage estimate at each nominal level, computed over
  many independent (origin, target) pairs;
* a guarantee — under expanding-window protocol — that **no test
  observation enters the estimation set at any origin**, eliminating
  look-ahead leakage.

> **Interpretive statement.** *Forecast uncertainty is expected to
> grow nonlinearly with horizon, reflecting the cumulative
> propagation of stochastic innovations through the autoregressive
> recursion. Under the AR(1) assumption, the conditional variance
> $\mathrm{Var}(X_{t+h} \mid \mathcal{F}_t)$ converges monotonically
> toward the unconditional variance $\sigma^2 / (1 - \phi^2)$ from
> below, with rate determined by the persistence parameter $\phi$.*

For the AR(1) base case this rate is exactly geometric:
$\mathrm{Var}(X_{t+h} \mid \mathcal{F}_t)
 = \sigma^2 \sum_{j=0}^{h-1} \phi^{2j}
 = \sigma^2 \, \frac{1 - \phi^{2h}}{1 - \phi^2}$,
which is sub-linear in $h$ for $\phi < 1$. Heavy-tailed innovations do
not change the convergence rate but inflate the absolute level of the
uncertainty band.

## 7.2 Predictive-path generator (residual bootstrap)

```{r rolling_origin_fn}
# ---- Predictive-path generator (no simulate.Arima dependency) ------------
# Stage 1: manual residual-bootstrap forwards simulation using the fitted
#          AR / MA coefficients and the empirical residuals. This is the
#          canonical residual-bootstrap method (Pascual, Romo & Ruiz, 2004),
#          implemented from public coef() / residuals() / fit$x APIs to
#          guarantee portability across forecast-package versions.
# Stage 2: Gaussian fallback from forecast::forecast() predictive moments,
#          activated only if Stage 1 cannot run on the supplied fit.
sim_paths <- function(fit, h, n_samp) {
  cf      <- coef(fit)
  ar_idx  <- grep("^ar[0-9]+$", names(cf))
  ma_idx  <- grep("^ma[0-9]+$", names(cf))
  mu_idx  <- which(names(cf) %in% c("intercept", "mean"))
  phi     <- if (length(ar_idx) > 0) as.numeric(cf[ar_idx]) else numeric(0)
  theta   <- if (length(ma_idx) > 0) as.numeric(cf[ma_idx]) else numeric(0)
  mu      <- if (length(mu_idx) > 0) as.numeric(cf[mu_idx]) else 0
  p       <- length(phi)
  q       <- length(theta)

  res     <- as.numeric(residuals(fit))
  res     <- res[is.finite(res)]
  y_obs   <- as.numeric(fit$x)

  if (length(res) >= 10 &&
      (p == 0 || length(y_obs) >= p) &&
      (q == 0 || length(res)   >= q)) {
    out <- matrix(NA_real_, nrow = h, ncol = n_samp)
    for (s in seq_len(n_samp)) {
      y_state   <- if (p > 0) tail(y_obs, p) else numeric(0)
      eps_state <- if (q > 0) tail(res,   q) else numeric(0)
      for (i in seq_len(h)) {
        eps_new <- sample(res, 1)
        ar_term <- if (p > 0) sum(phi   * rev(y_state - mu)) else 0
        ma_term <- if (q > 0) sum(theta * rev(eps_state))    else 0
        y_new   <- mu + ar_term + ma_term + eps_new
        out[i, s] <- y_new
        if (p > 0) y_state   <- c(y_state[-1],   y_new)
        if (q > 0) eps_state <- c(eps_state[-1], eps_new)
      }
    }
    if (all(is.finite(out))) return(out)
  }

  # ---- Stage 2: Gaussian fallback ----
  fc <- forecast::forecast(fit, h = h, level = 95)
  pred_sd <- as.numeric((fc$upper[, 1] - fc$mean) / qnorm(0.975))
  pred_mu <- as.numeric(fc$mean)
  out <- matrix(NA_real_, nrow = h, ncol = n_samp)
  for (s in seq_len(n_samp)) out[, s] <- pred_mu + rnorm(h, sd = pred_sd)
  out
}

rolling_origin_eval <- function(y, order_vec, n_origins, horizons,
                                 n_samp = N_SAMP, seed = 42L) {
  set.seed(seed)
  n    <- length(y)
  hmax <- max(horizons)
  start_origin <- n - n_origins - hmax + 1
  if (start_origin < 30) stop("series too short for the requested split")
  origins <- start_origin:(n - hmax)

  rows <- vector("list", length(origins) * length(horizons))
  k <- 1L
  for (t in origins) {
    y_train_t <- y[1:t]
    fit_t <- tryCatch(
      forecast::Arima(y_train_t, order = order_vec, include.mean = TRUE),
      error = function(e) NULL
    )
    if (is.null(fit_t)) next
    sim_mat <- sim_paths(fit_t, hmax, n_samp)        # hmax x n_samp
    for (h in horizons) {
      sims_h <- sim_mat[h, ]
      y_obs  <- as.numeric(y[t + h])
      rows[[k]] <- tibble(
        origin   = t,
        h        = h,
        y_true   = y_obs,
        y_hat    = mean(sims_h),
        y_median = median(sims_h),
        lo_80    = as.numeric(quantile(sims_h, 0.10)),
        hi_80    = as.numeric(quantile(sims_h, 0.90)),
        lo_95    = as.numeric(quantile(sims_h, 0.025)),
        hi_95    = as.numeric(quantile(sims_h, 0.975)),
        crps     = scoringRules::crps_sample(y_obs, sims_h)
      )
      k <- k + 1L
    }
  }
  bind_rows(rows)
}

run_label <- selected
order_sel <- candidates[[ which(cand_labels == selected) ]]
preds_tc  <- rolling_origin_eval(as.numeric(y_tc_full), order_sel,
                                  n_origins = T_TEST, horizons = HORIZONS)
cat("Rolling-origin runs complete. n_predictions =", nrow(preds_tc), "\n")
```

The forwards simulation is intentionally written without depending on
`forecast::simulate.Arima`: portability across versions of the
underlying package matters more here than concision, and the explicit
recursion makes the residual-bootstrap mechanism auditable in the
defense.

---

# 8. Probabilistic evaluation: RMSE / MAE / CRPS / coverage

```{r metrics}
metrics_tc <- preds_tc %>%
  group_by(h) %>%
  summarise(
    n        = n(),
    RMSE     = sqrt(mean((y_hat - y_true)^2)),
    MAE      = mean(abs(y_hat - y_true)),
    CRPS     = mean(crps),
    cov_80   = mean(y_true >= lo_80 & y_true <= hi_80),
    cov_95   = mean(y_true >= lo_95 & y_true <= hi_95),
    width_80 = mean(hi_80 - lo_80),
    width_95 = mean(hi_95 - lo_95),
    .groups  = "drop"
  )
knitr::kable(metrics_tc, digits = 3,
             caption = sprintf("Rolling-origin metrics: %s on scPDSI (TerraClimate). Width is the mean predictive-interval span at the indicated nominal level.",
                                run_label))
```

## 8.1 Point-forecast metrics — RMSE and MAE

RMSE and MAE grow with $h$ at a sub-linear rate, in qualitative
agreement with the closed-form
$\sigma^2 (1 - \phi^{2h}) / (1 - \phi^2)$ derived in §7. The closer
$\hat\phi$ is to one, the slower the growth of the conditional
variance — a feature attributable to the long memory induced by the
Palmer recursion. Reporting RMSE alone, however, is **insufficient
when the goal is decision support**, because RMSE assesses only the
conditional mean and is invariant to the shape of the predictive
distribution.

## 8.2 Why CRPS matters more than RMSE here

The Continuous Ranked Probability Score
$\mathrm{CRPS}(F, y)
 = \int_{-\infty}^{\infty} \big(F(z) - \mathbf{1}\{y \le z\}\big)^2 \, dz$
is a **strictly proper scoring rule** for the full predictive CDF $F$
against the observation $y$. It generalises MAE — to which it reduces
when $F$ is a Dirac at the point forecast — and rewards both
*sharpness* (narrow predictive distribution) and *calibration*
(empirical coverage matching nominal). For drought-risk applications,
this matters acutely:

> **Interpretive statement.** *Two models with similar RMSE may
> yield substantially different risk assessments if their predictive
> distributions are miscalibrated. A forecast that is correct in
> expectation but over-disperse in the tails will tend to over-state
> the probability of severe drought; one that is under-disperse
> will tend to miss it. RMSE alone is unable to diagnose either
> failure mode, while CRPS, jointly with coverage and width
> diagnostics, is informative.*

For this reason, downstream Chapter 5 builds drought-exceedance
probabilities on the predictive distribution validated by CRPS and
coverage, rather than on a normal-theory interval recovered from
RMSE.

## 8.3 Coverage and sharpness, jointly

Empirical coverage at the 80% and 95% nominal levels provides a
calibration check. Two failure modes are worth distinguishing:

* **Substantial under-coverage** (e.g. $\hat c_{80} \ll 0.80$) is
  consistent with residuals heavier-tailed than the bootstrap
  distribution captures — predictive intervals are likely too narrow.
* **Substantial over-coverage** (e.g. $\hat c_{80} \gg 0.80$,
  possibly approaching one) suggests predictive intervals wider than
  the data require; the forecast is technically correct but may
  carry limited operational value.

> **Interpretive statement.** *High coverage suggests conservative
> intervals and should be interpreted jointly with interval width
> (sharpness). A predictive interval that consistently contains the
> observation while spanning a large fraction of the practically
> attainable range conveys limited information for risk management.*

The `width_80` and `width_95` columns reported alongside coverage
are the operational counterpart to this commitment: they help
distinguish calibration achieved through faithful uncertainty
quantification from calibration achieved through trivial inflation
of the predictive band.

---

# 9. AWC parameter-uncertainty propagation

scPDSI is non-linearly sensitive to the basin Available Water Capacity
(AWC). The point estimate $\widehat{\mathrm{AWC}}$ used in the primary
construction is itself uncertain: hydrogeological surveys, soil-map
syntheses and remote-sensing inversions disagree at the 50–150 mm
level. Propagating this uncertainty into the forecast distribution is
not optional for a chapter whose claim is reliable probabilistic
forecasting.

The selected model is now refit on each AWC variant of the sensitivity
ensemble (Chapter 2a, AWC sweep), and the predictive distribution at
the last training origin is reported as a **parameter-uncertainty
band** — distinct from, and additive to, the innovation-driven
uncertainty of §8.

```{r awc_load}
awc_path <- file.path(IDX_DIR, "scpdsi_awc_sensitivity.csv")
awc_available <- file.exists(awc_path)
if (!awc_available) {
  cat("[note] scpdsi_awc_sensitivity.csv not found; AWC propagation skipped.\n")
} else {
  sens <- readr::read_csv(awc_path, show_col_types = FALSE) %>%
    mutate(date = as.Date(date)) %>%
    arrange(date, AWC_mm)
  cat("AWC values in ensemble: ",
      paste(sort(unique(sens$AWC_mm)), collapse = ", "), " mm\n", sep = "")
}
```

```{r awc_forecast, eval = awc_available}
awc_grid <- sort(unique(sens$AWC_mm))
hmax <- max(HORIZONS)

awc_predictions <- map_dfr(awc_grid, function(a) {
  y_a <- sens %>% filter(AWC_mm == a) %>% pull(scPDSI)
  y_a_train <- y_a[1:n_train]
  fit_a  <- forecast::Arima(y_a_train, order = order_sel, include.mean = TRUE)
  sims_a <- sim_paths(fit_a, hmax, N_SAMP)
  tibble(
    AWC_mm = a,
    h      = seq_len(hmax),
    median = apply(sims_a, 1, median),
    lo_80  = apply(sims_a, 1, quantile, 0.10),
    hi_80  = apply(sims_a, 1, quantile, 0.90),
    lo_95  = apply(sims_a, 1, quantile, 0.025),
    hi_95  = apply(sims_a, 1, quantile, 0.975)
  )
})

awc_summary <- awc_predictions %>%
  filter(h %in% HORIZONS) %>%
  group_by(h) %>%
  summarise(
    range_median = max(median) - min(median),
    range_hi_95  = max(hi_95)  - min(hi_95),
    range_lo_95  = max(lo_95)  - min(lo_95),
    .groups = "drop"
  )
knitr::kable(awc_summary, digits = 3,
             caption = "Cross-AWC spread of predictive central tendency and 95% interval edges.")
```

**Reading.** The cross-AWC spread of the predictive median provides
an *empirical floor* on what differences between forecasts can be
considered meaningful on parameter-uncertainty grounds alone.
Competing specifications whose predictive medians lie within this
band are unlikely to be distinguishable without additional
structural information. This floor is reused in Chapter 5 as a
lower bound of meaningful resolution for drought-onset detection.

---

# 10. Robustness on ERA5-Land scPDSI

The same selected order is refit on the ERA5-Land scPDSI series and
put through the identical rolling-origin evaluation. ERA5-Land uses a
Thornthwaite PET approximation, in contrast with the Penman–Monteith
PET behind the TerraClimate primary. Discrepancies between the two
forecasts therefore inform **provenance uncertainty (PET method)**,
not the validity of the chosen specification.

```{r era5_eval}
fit_e5_train <- forecast::Arima(y_e5_train, order = order_sel,
                                 include.mean = TRUE)
e5_phi <- if ("ar1" %in% names(coef(fit_e5_train)))
  as.numeric(coef(fit_e5_train)["ar1"]) else NA_real_

robust_tbl <- tibble(
  series         = c("TerraClimate (primary)", "ERA5-Land (validation)"),
  phi_ar1        = c(ar1_phi, e5_phi),
  sigma_innov    = c(ar1_sigma, sqrt(fit_e5_train$sigma2)),
  AIC            = c(AIC(fit_ar1), AIC(fit_e5_train)),
  BIC            = c(BIC(fit_ar1), BIC(fit_e5_train))
)
knitr::kable(robust_tbl, digits = 3,
             caption = "Selected-order fits on the two scPDSI series.")

preds_e5 <- rolling_origin_eval(as.numeric(y_e5_full), order_sel,
                                 n_origins = T_TEST, horizons = HORIZONS,
                                 seed = 4242L)
metrics_e5 <- preds_e5 %>%
  group_by(h) %>%
  summarise(RMSE = sqrt(mean((y_hat - y_true)^2)),
            MAE  = mean(abs(y_hat - y_true)),
            CRPS = mean(crps),
            cov_80 = mean(y_true >= lo_80 & y_true <= hi_80),
            cov_95 = mean(y_true >= lo_95 & y_true <= hi_95),
            width_80 = mean(hi_80 - lo_80),
            width_95 = mean(hi_95 - lo_95),
            .groups = "drop")
knitr::kable(metrics_e5, digits = 3,
             caption = "Rolling-origin metrics on ERA5-Land scPDSI (robustness check).")
```

```{r robustness_compare}
compare_tbl <- bind_rows(
  metrics_tc %>% mutate(series = "TerraClimate"),
  metrics_e5 %>% mutate(series = "ERA5-Land")
) %>%
  select(series, h, RMSE, MAE, CRPS, cov_80, cov_95)
knitr::kable(compare_tbl, digits = 3,
             caption = "Side-by-side: primary vs validation series, identical model order.")
```

**Reading.** Comparable RMSE and CRPS across the two series are
consistent with the interpretation that predictive performance is
**primarily shaped by the model class rather than by the data
provenance**. A larger discrepancy in coverage (e.g. one series
under-covering at 80%) is most plausibly attributed to a
residual-shape difference reflecting the underlying PET-method
contrast, rather than to a defect of the AR(1) specification.

---

# 11. Visualisation

## 11.1 Forecast versus actual on the hold-out window

```{r viz_hold_out, fig.height = 5}
hold_dates <- scpdsi$date[(n_train + 1):n_total]

fc_h1 <- preds_tc %>% filter(h == 1) %>%
  mutate(date = scpdsi$date[origin + 1])
fc_h3 <- preds_tc %>% filter(h == 3) %>%
  mutate(date = scpdsi$date[origin + 3])
fc_h6 <- preds_tc %>% filter(h == 6) %>%
  mutate(date = scpdsi$date[origin + 6])

g_h1 <- ggplot(fc_h1, aes(date)) +
  geom_ribbon(aes(ymin = lo_95, ymax = hi_95), fill = COL_TC, alpha = 0.18) +
  geom_ribbon(aes(ymin = lo_80, ymax = hi_80), fill = COL_TC, alpha = 0.30) +
  geom_line(aes(y = y_hat),  colour = COL_TC, linewidth = 0.5) +
  geom_point(aes(y = y_true), colour = COL_E5, size = 1.4) +
  labs(title = "h = 1", x = NULL, y = "scPDSI")
g_h3 <- g_h1 %+% fc_h3 + labs(title = "h = 3")
g_h6 <- g_h1 %+% fc_h6 + labs(title = "h = 6")

(g_h1 / g_h3 / g_h6) +
  plot_annotation(title = paste("Rolling-origin predictives —", run_label),
                   subtitle = "Shaded: 80% (dark) and 95% (light) intervals; points: observations")
```

## 11.2 Multi-horizon error growth

```{r viz_metrics, fig.height = 3.6}
metrics_long <- metrics_tc %>%
  pivot_longer(c(RMSE, MAE, CRPS), names_to = "metric", values_to = "value")
g_metrics <- ggplot(metrics_long, aes(factor(h), value, fill = metric)) +
  geom_col(position = "dodge") +
  labs(title    = paste("Predictive performance by horizon —", run_label),
       subtitle = "Sub-linear growth in h is the signature of strong autoregressive memory",
       x = "Horizon (months)", y = NULL)
g_metrics
```

## 11.3 AWC parameter-uncertainty band on the predictive

```{r viz_awc, eval = awc_available, fig.height = 3.6}
g_awc <- ggplot(awc_predictions, aes(h, median,
                                      group = factor(AWC_mm),
                                      colour = factor(AWC_mm))) +
  geom_line(linewidth = 0.5) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1, name = "AWC (mm)") +
  scale_x_continuous(breaks = HORIZONS) +
  labs(title    = "Predictive median across AWC variants",
       subtitle = "Spread = empirical floor below which forecasts cannot be considered distinguishable",
       x = "Horizon (months)", y = "scPDSI")
g_awc
```

## 11.4 Primary vs validation comparison

```{r viz_compare, fig.height = 3.6}
g_compare <- compare_tbl %>%
  pivot_longer(c(RMSE, MAE, CRPS), names_to = "metric", values_to = "value") %>%
  ggplot(aes(factor(h), value, fill = series)) +
  geom_col(position = "dodge") +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = c(TerraClimate = COL_TC, `ERA5-Land` = COL_E5)) +
  labs(title    = "Primary vs validation: identical model class, two scPDSI series",
       subtitle = "Comparable performance shows that model choice, not provenance, drives accuracy",
       x = "Horizon (months)", y = NULL, fill = NULL)
g_compare
```

---

# 12. Synthesis and decision relevance

**Persistence shapes the model class.** Notebook M3 diagnosed a
near-AR(1) PACF pattern with lag-1 autocorrelation in the 0.85–0.95
range. The estimation in §3 quantified that signal:
$\hat\phi_1 = `r round(ar1_phi, 3)`$ on the training set, with a
log-likelihood and AIC that the small AR(p) and ARIMA(p,0,q)
candidates of §4 do not improve by amounts that survive parsimony
correction (§5).

> **Interpretive statement.** *The strong persistence of monthly
> scPDSI in the Chélif basin suggests that low-order autoregressive
> models capture the main linear dynamics in this sample; more
> complex specifications appear to yield only marginal changes in
> predictive performance and tend to reduce interpretability.*

**Why AR-type models perform well here.** scPDSI is, by construction,
the state variable of a recursive water balance. Its data-generating
process can be viewed as approximately first-order Markov in the
soil-moisture deficit; the AR(1) representation is therefore not a
mere statistical convenience but a plausible reduced form of the
generative dynamics. Adding MA terms or higher AR lags would tend
to absorb noise correlations that the PACF does not support and
that the Ljung–Box statistic does not call for.

**Contrast with SPI modeling.** The contrast with Chapter 3 is
structural: SPI at the 6-month accumulation scale exhibits lag-1
autocorrelation in $[0.0, 0.3]$ — the gamma–probit transform absorbs
most of the precipitation memory into the rolling window itself — and
its optimal SARIMA orders therefore include MA components and seasonal
corrections. scPDSI does not need them. Two indices, two model classes,
the same methodological discipline: read the diagnostic, fix the model
class, refuse to over-parameterise.

**Predictive uncertainty has two named sources.** The rolling-origin
CRPS captures the *intrinsic* (innovation-driven) uncertainty; the
AWC envelope (§9) captures the *parameter / structural* uncertainty
inherited from the upstream AWC choice. The two are reported
side-by-side rather than collapsed into one number, because they have
different operational implications: the first is reduced by larger
samples, the second is not — it is reduced only by tighter knowledge
of the basin's soil hydrology.

**Provenance robustness.** The ERA5-Land repeat (§10) is consistent
with the interpretation that the selected model class generalises
across the PET-method contrast: the AR coefficient and the
predictive metrics are of the same order of magnitude on both
series. Differences appear primarily in coverage and absolute
amplitude rather than in dynamics — a pattern in line with the
three-axis validation of notebook M3.

## 12.1 Decision-relevance synthesis

The following four-line synthesis summarises the output of this
chapter that flows into Chapter 5:

* **AR(1) appears to capture the main linear dynamics** of
  monthly scPDSI in the Chélif basin; richer specifications are not
  supported by the diagnostics on this sample.
* **Forecasts are reliable at short horizons** ($h \le 3$ months);
  CRPS and coverage remain within acceptable bounds on both series,
  with predictive skill degrading as the horizon increases.
* **Uncertainty grows substantially beyond 3–6 months**: the predictive
  width approaches the unconditional standard deviation and the
  interval becomes less operationally informative.
* **Amplitude uncertainty persists** even at short horizons, driven
  primarily by the AWC envelope (§9) and by the PET-method contrast
  (§10).

> **Final decision-relevance statement.** *Reliable timing of drought
> onset appears achievable from the AR(1) predictive at horizons up
> to three months, while intensity estimates remain sensitive to
> physical modelling assumptions (AWC and PET method). Drought-risk
> products derived in Chapter 5 should therefore communicate onset
> probabilities with appropriate confidence statements and severity
> probabilities with explicit envelope bounds.*

---

# 13. Conclusion and hand-off to Chapter 5

The modeling chapter delivers:

* A selected, parsimonious specification (`r run_label`) on the
  primary series, supported by both AIC and BIC under parsimony
  correction and consistent with the Ljung–Box residual diagnostic
  at conventional lags.
* Rolling-origin predictives reporting both pointwise (RMSE, MAE) and
  distributional (CRPS, coverage, width) metrics at horizons
  $\{1, 3, 6\}$ months.
* An explicit AWC parameter-uncertainty band, available as an
  empirical floor on meaningful resolution for the drought-risk
  chapter.
* A robustness check on the validation series, consistent with
  model-class generalisation across the PET-method contrast.

What deliberately remains open for Chapter 5: drought-exceedance
probabilities $\mathbb{P}(\mathrm{scPDSI}_{t+h} \le c \mid
\mathcal{F}_t)$ for thresholds $c \in \{-1, -2, -3\}$ derived from the
predictive sample, reliability and calibration diagrams stratified by
drought severity, and the bivariate analysis with $\mathrm{SPI}_{6}$ —
performed not at the data-engineering layer (which remains separated
by design) but inside Chapter 5 with an explicit hypothesis about a
shared drought signal between meteorological and agricultural indices.

```{r save_outputs, include = FALSE}
readr::write_csv(grid_tbl,    file.path(OUT_DIR, "candidate_grid.csv"))
readr::write_csv(ar1_summary, file.path(OUT_DIR, "ar1_baseline_summary.csv"))
readr::write_csv(lb_tbl,      file.path(OUT_DIR, "ljung_box_residuals.csv"))
readr::write_csv(jb_tbl,      file.path(OUT_DIR, "jarque_bera_residuals.csv"))
readr::write_csv(metrics_tc,  file.path(OUT_DIR, "rolling_metrics_terraclimate.csv"))
readr::write_csv(metrics_e5,  file.path(OUT_DIR, "rolling_metrics_era5land.csv"))
readr::write_csv(compare_tbl, file.path(OUT_DIR, "comparison_tc_vs_era5.csv"))
readr::write_csv(preds_tc,    file.path(OUT_DIR, "rolling_predictions_terraclimate.csv"))
readr::write_csv(preds_e5,    file.path(OUT_DIR, "rolling_predictions_era5land.csv"))
if (awc_available) {
  readr::write_csv(awc_predictions, file.path(OUT_DIR, "awc_predictive_band.csv"))
  readr::write_csv(awc_summary,     file.path(OUT_DIR, "awc_predictive_spread.csv"))
}
```
