---
title: "M3 — scPDSI Diagnostics and Validation (Supplementary)"
subtitle: "Drought Modeling and Probabilistic Forecasting in the Chélif Basin (Algeria)"
author: |
  Imene DERRAR  
  Supervisor: Dr. Abdelillah OTMANE CHERIF  
  NHSM (National Higher School of Mathematics) · IHFR (Institut Hydrométéorologique de Formation et de Recherche)  
  Final-year engineering thesis in Applied Mathematics — Academic year 2025–2026
date: ""
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
    theme: readable
    highlight: tango
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width = 9,
  fig.height = 4.5,
  fig.align = "center",
  dpi       = 110,
  cache     = FALSE
)
```

# Role of this notebook

This notebook is a **supplementary diagnostic and validation companion** to
the scPDSI track of the thesis. The raw computation is performed by the
scPDSI pipeline (notebook E.8) and the scPDSI compute step (notebook E.7);
this notebook does **not** recompute the index. It operates on the
persisted artefacts (`scpdsi_master.csv`, `scpdsi_awc_sensitivity.csv`,
`scpdsi_meta.json`) to provide five diagnostic checks that support the
interpretation given in the thesis Chapter 4 and the historical-episode
discussion:

1. **Structural alignment.** Verify that the scPDSI monthly index matches
   the SPI master index exactly (1990-01 → 2015-12, 312 observations).
2. **Cross-validation under three axes.** Decompose the TerraClimate-vs-ERA5
   agreement into *temporal*, *amplitude* and *categorical* components, each
   addressing a distinct operational question.
3. **Formal stochastic characterisation.** Formal stationarity protocol
   (ADF + KPSS), ACF/PACF with quantitative interpretation of the decay,
   and residual diagnostics from a baseline AR(1) fit (Ljung–Box,
   Jarque–Bera).
4. **Parameter-uncertainty propagation.** Quantify the AWC sensitivity
   envelope (range, variance, episode-level deformation), not merely
   visualise it.
5. **Modeling-side handoff.** Justify, on the basis of those diagnostics, the
   model class for Chapter 4 (autoregressive, with explicit memory) and the
   convention that SPI and scPDSI remain *separate stochastic processes* at
   the data-engineering layer.

The notebook follows the same structure as the SPI diagnostics:

1. Experimental setup (paths, seed, hyper-parameters)
2. Data loading and structural validation
3. Exploratory analysis (descriptive, distributional shape, time series)
4. Validation: TerraClimate vs ERA5-Land — three-axis decomposition
5. Formal stationarity testing (ADF + KPSS, joint reading)
6. Time-series structure (ACF/PACF, quantitative)
7. Residual diagnostics from a baseline AR(1) fit
8. AWC parameter-uncertainty propagation
9. Visualisation integration with the canonical figure set
10. Transition to the modeling phase (Chapter 4)
11. Conclusion

---

# 1. Experimental setup

```{r packages}
# ---- 1.1  Packages -------------------------------------------------------
# Each package is included for a specific function:
#   tidyverse / lubridate : data wrangling, dates, plotting (ggplot2)
#   zoo                   : rolling-window operations
#   forecast              : ggAcf, ggPacf, Arima(), residual extraction
#   tseries, urca         : adf.test, kpss.test, jarque.bera.test
#   patchwork             : ggplot panel composition
#   knitr / kableExtra    : presentable tables in the rendered notebook
#   jsonlite              : read scpdsi_meta.json provenance
suppressPackageStartupMessages({
  pkgs <- c("tidyverse", "lubridate", "zoo",
            "forecast", "tseries", "urca",
            "patchwork", "knitr", "kableExtra", "jsonlite")
  for (p in pkgs) {
    if (!requireNamespace(p, quietly = TRUE)) {
      install.packages(p, repos = "https://cloud.r-project.org")
    }
    library(p, character.only = TRUE)
  }
})

# ---- 1.2  Publication theme ---------------------------------------------
# A single thesis-wide ggplot theme so every figure renders identically.
THESIS_THEME <- theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 10, colour = "grey25"),
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    strip.text       = element_text(face = "bold", size = 10),
    legend.position  = "bottom",
    legend.title     = element_text(size = 9),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
theme_set(THESIS_THEME)

# Palette — TerraClimate / ERA5-Land carry consistent colours throughout.
COL_TC  <- "#1F3864"   # primary
COL_E5  <- "#C00000"   # validation
COL_GREY <- "grey45"

options(scipen = 6, digits = 4)
```

```{r config}
# ---- 1.3  Project paths, seed, global parameters --------------------------
# All scPDSI artifacts live under _structured_project/ after the May 2026
# methodological pivot (shared station precipitation for SPI and scPDSI).
# Raw NetCDFs and the basin shapefile remain in LEGACY_ROOT/01_data_raw/
# but the analysis notebook never touches those directly.
LEGACY_ROOT     <- Sys.getenv("CHELIF_PROJECT_ROOT", unset = getwd())
STRUCTURED_ROOT <- file.path(LEGACY_ROOT, "_structured_project")
BASE_DIR        <- STRUCTURED_ROOT

IDX_DIR  <- file.path(BASE_DIR, "02_processed", "drought_indices")
SPI_DIR  <- file.path(BASE_DIR, "04_outputs",
                      "02_spi_construction_and_validation")
FIG_DIR  <- file.path(BASE_DIR, "04_outputs",
                      "M3_scpdsi_diagnostics", "figures")
OUT_DIR  <- file.path(BASE_DIR, "03_outputs_R", "M3_scpdsi_diagnostics")
if (!dir.exists(OUT_DIR)) dir.create(OUT_DIR, recursive = TRUE)

set.seed(42)

# Persisted artefacts (already computed) -----------------------------------
SCPDSI_MASTER     <- file.path(IDX_DIR, "scpdsi_master.csv")
SCPDSI_AWC_SWEEP  <- file.path(IDX_DIR, "scpdsi_awc_sensitivity.csv")
SCPDSI_META_JSON  <- file.path(IDX_DIR, "scpdsi_meta.json")
SPI_MASTER        <- file.path(SPI_DIR, "spi_regional_series.csv")

# Modeling-side conventions inherited from the SPI notebook ---------------
PRIMARY_SCPDSI_SRC    <- "TerraClimate"   # Penman-Monteith PET => primary
VALIDATION_SCPDSI_SRC <- "ERA5-Land"      # Thornthwaite PET => robustness

# Palmer drought thresholds (Palmer, 1965; Wells et al., 2004) ------------
PALMER_THR <- c(D0 = -1.0, D1 = -2.0, D2 = -3.0, D3 = -4.0)

# Methodology-chapter acceptance thresholds (§7) ---------------------------
TARGET_PEARSON   <- 0.85
TARGET_RMSE      <- 0.60
TARGET_CLASS_AGR <- 0.75
TARGET_HIT_D2    <- 0.80

cat("Configuration locked.\n",
    "  primary scPDSI    :", PRIMARY_SCPDSI_SRC, "\n",
    "  validation scPDSI :", VALIDATION_SCPDSI_SRC, "\n",
    "  Palmer thresholds :", paste(names(PALMER_THR), PALMER_THR,
                                   sep = "=", collapse = ", "), "\n",
    "  Targets           : r>=", TARGET_PEARSON,
                          " RMSE<=", TARGET_RMSE,
                          " class>=", TARGET_CLASS_AGR,
                          " hit(D2+)>=", TARGET_HIT_D2, "\n")
```

---

# 2. Data loading and structural validation

```{r load_data}
# ---- 2.1  Canonical SPI index -------------------------------------------
# The thesis-wide convention is that the SPI master CSV defines the time axis;
# scPDSI is expected to share the same axis. The stopifnot() below
# enforces this alignment at load time.
spi_raw <- readr::read_csv(SPI_MASTER, show_col_types = FALSE) %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)
SPI_INDEX <- spi_raw$date
SPI_START <- min(SPI_INDEX)
SPI_END   <- max(SPI_INDEX)

# ---- 2.2  scPDSI master and identity assertion --------------------------
scpdsi <- readr::read_csv(SCPDSI_MASTER, show_col_types = FALSE) %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)
stopifnot(identical(scpdsi$date, SPI_INDEX))

# ---- 2.3  Provenance metadata -------------------------------------------
meta <- jsonlite::fromJSON(SCPDSI_META_JSON, simplifyDataFrame = FALSE)

cat("scPDSI master loaded:\n",
    "  rows               :", nrow(scpdsi), "\n",
    "  range              :", as.character(SPI_START), "->",
                                as.character(SPI_END), "\n",
    "  primary PET method :", meta$datasets$primary$pet_method,    "\n",
    "  ERA5 PET method    :", meta$datasets$validation$pet_method, "\n",
    "  AWC central        :", meta$awc$central_mm, "mm\n",
    "  AWC sweep          :", paste(meta$awc$sensitivity_grid, collapse = ", "), "\n")
```

The `stopifnot(identical(...))` line is the contractual interface between this
notebook and the data-engineering layer: scPDSI is **not** allowed onto the
analysis bench unless its monthly index matches the SPI index exactly.
This eliminates the silent-misalignment risk that a fuzzy join could otherwise
mask.

---

# 3. Exploratory analysis

## 3.1 Descriptive statistics

```{r descriptive}
desc <- scpdsi %>%
  pivot_longer(c(scPDSI_TerraClimate, scPDSI_ERA5),
               names_to = "source", values_to = "x") %>%
  group_by(source) %>%
  summarise(
    n           = sum(!is.na(x)),
    mean        = mean(x,  na.rm = TRUE),
    sd          = sd(x,    na.rm = TRUE),
    min         = min(x,   na.rm = TRUE),
    p25         = quantile(x, 0.25, na.rm = TRUE),
    median      = median(x, na.rm = TRUE),
    p75         = quantile(x, 0.75, na.rm = TRUE),
    max         = max(x,   na.rm = TRUE),
    skewness    = mean((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) /
                   sd(x, na.rm = TRUE)^3,
    ex_kurtosis = mean((x - mean(x, na.rm = TRUE))^4, na.rm = TRUE) /
                   sd(x, na.rm = TRUE)^4 - 3,
    .groups = "drop"
  )
knitr::kable(desc, digits = 3,
             caption = "Descriptive statistics of the two scPDSI series (full window).")
```

**Interpretation.** Unlike SPI (constructed to be marginally $N(0,1)$),
scPDSI has no enforced unit-variance scale: its dispersion is governed by the
basin water-balance recursion and the self-calibration step (Wells et al.,
2004). A small mean offset between the two sources is therefore *not* a
defect — it reflects the asymmetric demand profiles imposed by Penman–Monteith
(TerraClimate) vs Thornthwaite (ERA5-Land). Heavy negative tails (negative
skewness) are the hydrologic signature expected for a water-balance index in
a semi-arid Mediterranean basin. Note that the ERA5-Land series mildly
breaches the conventional Palmer upper bound (sample max $\approx +5.6$ vs
the design ceiling of $+4$): on a 26-year self-calibration window, the
systematic under-estimation of summer PET by Thornthwaite leaves the wet end
of the distribution less constrained, producing an asymmetric upper tail not
present in the TerraClimate variant. This is a known limitation of
self-calibrated Palmer indices on short records and is bounded by the AWC
sensitivity envelope analyzed in §8.

## 3.2 Time series

```{r ts_plot, fig.height = 4.8}
g_ts <- scpdsi %>%
  pivot_longer(c(scPDSI_TerraClimate, scPDSI_ERA5),
               names_to = "source", values_to = "x") %>%
  mutate(source = recode(source,
                         scPDSI_TerraClimate = "TerraClimate (primary)",
                         scPDSI_ERA5         = "ERA5-Land (validation)")) %>%
  ggplot(aes(date, x, colour = source)) +
  annotate("rect", xmin = SPI_START, xmax = SPI_END,
           ymin = PALMER_THR["D1"], ymax = PALMER_THR["D0"],
           alpha = 0.10, fill = "#E08E79") +
  annotate("rect", xmin = SPI_START, xmax = SPI_END,
           ymin = PALMER_THR["D2"], ymax = PALMER_THR["D1"],
           alpha = 0.15, fill = "#C0504D") +
  annotate("rect", xmin = SPI_START, xmax = SPI_END,
           ymin = -Inf, ymax = PALMER_THR["D2"],
           alpha = 0.18, fill = "#7E0000") +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.3, colour = COL_GREY) +
  geom_line(linewidth = 0.45) +
  scale_colour_manual(values = c("TerraClimate (primary)" = COL_TC,
                                  "ERA5-Land (validation)" = COL_E5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title    = "scPDSI for the Chélif basin: TerraClimate vs ERA5-Land",
       subtitle = "Shaded bands indicate Palmer drought severity (D0-D3)",
       x = NULL, y = "scPDSI", colour = NULL)
g_ts
```

## 3.3 Distributional shape

```{r densities, fig.height = 3.6}
g_dens <- scpdsi %>%
  pivot_longer(c(scPDSI_TerraClimate, scPDSI_ERA5),
               names_to = "source", values_to = "x") %>%
  mutate(source = recode(source,
                         scPDSI_TerraClimate = "TerraClimate",
                         scPDSI_ERA5         = "ERA5-Land")) %>%
  ggplot(aes(x, fill = source, colour = source)) +
  geom_density(alpha = 0.30, linewidth = 0.5) +
  geom_vline(xintercept = c(PALMER_THR, 0), linetype = 2,
             linewidth = 0.25, colour = COL_GREY) +
  scale_fill_manual(values   = c(TerraClimate = COL_TC, `ERA5-Land` = COL_E5)) +
  scale_colour_manual(values = c(TerraClimate = COL_TC, `ERA5-Land` = COL_E5)) +
  labs(title    = "Empirical densities of scPDSI",
       subtitle = "Vertical lines: Palmer thresholds (D0 to D3) and zero",
       x = "scPDSI", y = "Density", fill = NULL, colour = NULL)
g_dens
```

The densities are *bimodal-asymmetric* in this basin: a main lobe near
the climatological mean, a heavier left tail populated by drought episodes.
ERA5-Land sits slightly to the right of TerraClimate (less negative),
consistent with the smaller PET demand produced by Thornthwaite during
summer months — see §4.

---

# 4. Validation analysis: TerraClimate vs ERA5-Land

We decompose the TerraClimate / ERA5-Land agreement into **three orthogonal
axes**, because aggregate metrics in isolation are misleading: a series can
agree in shape while disagreeing in amplitude, or agree in central tendency
while disagreeing in classification near a threshold.

| axis | metric | operational question |
|---|---|---|
| Temporal      | Pearson $r$, lag-0   | Do the two indices follow the same drought trajectory? |
| Amplitude     | RMSE, mean residual  | How large is the typical pointwise disagreement? |
| Categorical   | Class agreement, D2+ hit rate | Do they identify the same Palmer category and the same severe episodes? |

```{r palmer_class_fn}
palmer_class <- function(x) {
  cut(x,
      breaks = c(-Inf, -4, -3, -2, -1, 1, 2, 3, 4, Inf),
      labels = c("Extreme drought", "Severe drought", "Moderate drought",
                 "Mild drought",  "Near normal",
                 "Mild wet", "Moderate wet", "Severe wet", "Extreme wet"),
      right  = FALSE)
}
```

## 4.1 Three-axis metrics

```{r metrics}
v <- scpdsi %>%
  mutate(diff      = scPDSI_TerraClimate - scPDSI_ERA5,
         class_tc  = palmer_class(scPDSI_TerraClimate),
         class_e5  = palmer_class(scPDSI_ERA5),
         severe_tc = scPDSI_TerraClimate <= -2,
         severe_e5 = scPDSI_ERA5 <= -2)

pearson_r       <- cor(v$scPDSI_TerraClimate, v$scPDSI_ERA5, use = "complete.obs")
rmse_idx        <- sqrt(mean(v$diff^2, na.rm = TRUE))
mean_diff       <- mean(v$diff, na.rm = TRUE)
class_agreement <- mean(v$class_tc == v$class_e5, na.rm = TRUE)
hit_rate_d2     <- mean(v$severe_tc == v$severe_e5, na.rm = TRUE)

# OLS slope: a slope < 1 means TerraClimate amplitude > ERA5 amplitude.
ols_fit <- lm(scPDSI_ERA5 ~ scPDSI_TerraClimate, data = v)
ols_slope     <- as.numeric(coef(ols_fit)["scPDSI_TerraClimate"])
ols_intercept <- as.numeric(coef(ols_fit)["(Intercept)"])

metrics_tbl <- tibble(
  axis      = c("Temporal", "Amplitude", "Amplitude",
                "Amplitude", "Categorical", "Categorical"),
  metric    = c("Pearson r",
                "RMSE (idx units)",
                "Mean residual (TC - ERA5)",
                "OLS slope (ERA5 ~ TC)",
                "Drought-class agreement (D0-D4)",
                "Hit rate on D2+ events"),
  value     = c(pearson_r, rmse_idx, mean_diff,
                ols_slope, class_agreement, hit_rate_d2),
  threshold = c(paste0(">= ", TARGET_PEARSON),
                paste0("<= ", TARGET_RMSE),
                "approx 0",
                "approx 1",
                paste0(">= ", TARGET_CLASS_AGR),
                paste0(">= ", TARGET_HIT_D2)),
  passes    = c(pearson_r        >= TARGET_PEARSON,
                rmse_idx         <= TARGET_RMSE,
                abs(mean_diff)   <= 0.30,
                abs(ols_slope-1) <= 0.20,
                class_agreement  >= TARGET_CLASS_AGR,
                hit_rate_d2      >= TARGET_HIT_D2)
)
knitr::kable(metrics_tbl, digits = 3,
             caption = "Validation metrics decomposed along three axes (temporal / amplitude / categorical).")
```

## 4.2 Reading the result: an isolated PET-method signature

The empirical results split cleanly along the three axes, and — because
precipitation is now identical between the two variants by construction
(Chapter 2.2) — every axis of disagreement is directly attributable to the
PET formulation:

* **Temporal axis — strong agreement.** Pearson $r = 0.895$ supports the reading that
  the two series describe the same drought *dynamics*: the timing, ordering
  and intensity ranking of episodes coincide across the 312-month record.
  The D2+ hit rate of $0.93$ corroborates this on the severe sub-domain —
  both variants identify the same major drought episodes.
* **Amplitude axis — systematic compression.** RMSE $= 1.009$ scPDSI units
  (above the $0.60$ acceptance threshold), an OLS slope of $0.834$, and a
  small but non-zero mean residual of $+0.106$ jointly indicate that
  ERA5-Land's amplitude is *compressed* relative to TerraClimate's by
  approximately $17\%$. The compression is in the dry direction: the same
  precipitation deficit produces a less severe scPDSI signal under
  Thornthwaite PET than under Penman–Monteith, because Thornthwaite reports
  a lower atmospheric demand and therefore a less depleted soil-moisture
  state.
* **Categorical axis — operational divergence.** Palmer-class agreement
  $= 0.47$ (below the $0.75$ threshold) signals that the two products place
  the same month in *different drought categories* roughly half of the time,
  despite their excellent inter-monthly correlation.

This is the textbook signature of two indices forced with **identical
precipitation** but driven by **different evapotranspiration formulations**,
observed without the confounding effect of disagreement in the precipitation
input:

| | TerraClimate | ERA5-Land |
|---|---|---|
| PET method | Modified Penman–Monteith (full energy balance) | Thornthwaite (1948), latitude-corrected |
| Inputs used | $T$, radiation, vapour pressure, wind | $T_{2m}$ only |
| Known bias in semi-arid summer | small | underestimates PET by $10$–$25\%$ |

Because both indices now ingest the same 9-station basin-mean precipitation
(Chapter 2.2), the divergence is no longer interpretable as a mix of
P-source and PET-method effects: it is a **clean measurement** of how much
the Penman–Monteith vs Thornthwaite contrast propagates through the
recursive Palmer water balance. The amplitude compression (slope $0.834$,
mean residual $+0.106$) is the propagation through the index *level*; the
categorical disagreement ($0.47$) is the propagation through the *Palmer
thresholds*.

> **Methodological note.** *Under the pre-pivot architecture (each variant
> consuming its product-internal precipitation and PET) this comparison
> reported $r = 0.935$ and class agreement $= 0.55$. The improvement was
> illusory: it reflected the internal consistency of each product with
> itself, not a measurement of the PET-method effect. The post-pivot
> $r = 0.895$ is a less optimistic but methodologically purer number, and
> the categorical agreement of $0.47$ is the true measurement of how often
> Thornthwaite places a marginal month in a different Palmer class than
> Penman–Monteith would.*

The implication for Chapter 4 is unchanged in direction but sharper in
interpretation: TerraClimate (Penman–Monteith) is the **primary** series
for inference and forecasting; ERA5-Land (Thornthwaite) is the
**lower bound** on PET-method uncertainty, not a competing measurement.
The categorical disagreement reported here is therefore an *uncertainty
envelope around the operational drought call* and should be
carried forward into Chapter 4's predictive distributions rather than
absorbed into them.
The D2+ hit rate of $0.93$ provides empirical support that the
*severe* operational signal — the part most relevant for water-management
decisions — is robust to PET-method choice, even though marginal
classifications are not.

## 4.3 Scatter and residual diagnostics

```{r scatter_resid, fig.height = 4.0}
g_scatter <- ggplot(v, aes(scPDSI_TerraClimate, scPDSI_ERA5)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = COL_GREY) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = COL_TC, linewidth = 0.6) +
  geom_point(alpha = 0.55, size = 1.0, colour = COL_TC) +
  coord_equal() +
  labs(title    = "TerraClimate vs ERA5-Land",
       subtitle = sprintf("Pearson r = %.3f; OLS slope = %.3f, intercept = %.3f",
                          pearson_r, ols_slope, ols_intercept),
       x = "scPDSI (TerraClimate)", y = "scPDSI (ERA5-Land)")

g_resid <- ggplot(v, aes(date, diff)) +
  geom_hline(yintercept = 0,         colour = COL_GREY) +
  geom_hline(yintercept = mean_diff, colour = COL_E5,
             linetype = 2, linewidth = 0.4) +
  geom_line(linewidth = 0.4, colour = COL_E5) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title    = "Residuals: TerraClimate minus ERA5-Land",
       subtitle = sprintf("RMSE = %.3f; mean residual = %.3f (red dashed)",
                          rmse_idx, mean_diff),
       x = NULL,
       y = expression(Delta * " scPDSI " == "TC - ERA5"))

g_scatter + g_resid + plot_layout(ncol = 2, widths = c(1, 1.1))
```

The OLS slope below 1 (visible as a tilt of the regression line away from the
identity reference) is the empirical signature of the Thornthwaite vs
Penman–Monteith mismatch: TerraClimate produces a wider scPDSI dispersion
because its higher summer PET demand widens the moisture-deficit signal in
dry years. The residual time series oscillates around a *non-zero* mean
$(\bar{\Delta} = `r round(mean_diff, 3)`)$ — a structural offset, not
white-noise disagreement.

## 4.4 Drought episodes (corrected run-length on the canonical date axis)

```{r episodes_fn}
# Run-length episodes computed on the canonical date index.
# Each episode's length is the number of *consecutive months* with x <= -1.
# We anchor start/end on actual `date` values, not on row positions, to
# eliminate any risk of off-by-one indexing.
episodes_of <- function(x, dates, threshold = -1, min_length = 3L) {
  stopifnot(length(x) == length(dates))
  flag <- !is.na(x) & x <= threshold
  if (!any(flag)) return(tibble(start = as.Date(character()),
                                end   = as.Date(character()),
                                length_months = integer()))
  rle_obj <- rle(flag)
  ends   <- cumsum(rle_obj$lengths)
  starts <- ends - rle_obj$lengths + 1L
  tibble(
    start         = dates[starts][rle_obj$values],
    end           = dates[ends][rle_obj$values],
    length_months = as.integer(rle_obj$lengths[rle_obj$values])
  ) %>% filter(length_months >= min_length)
}

ep_tc <- episodes_of(scpdsi$scPDSI_TerraClimate, scpdsi$date) %>%
  mutate(source = "TerraClimate")
ep_e5 <- episodes_of(scpdsi$scPDSI_ERA5,         scpdsi$date) %>%
  mutate(source = "ERA5-Land")
ep <- bind_rows(ep_tc, ep_e5)

ep_summary <- ep %>%
  group_by(source) %>%
  summarise(n_episodes        = n(),
            mean_length_mo    = mean(length_months),
            median_length_mo  = median(length_months),
            max_length_mo     = max(length_months),
            total_drought_mo  = sum(length_months),
            .groups = "drop")
knitr::kable(ep_summary, digits = 1,
             caption = "Drought-episode statistics (scPDSI <= -1, length >= 3 months).")
```

```{r episodes_fig, fig.height = 4.5}
g_episodes <- ggplot(ep,
       aes(xmin = start, xmax = end, ymin = 0, ymax = length_months,
           fill = source)) +
  geom_rect(alpha = 0.85) +
  scale_fill_manual(values = c(TerraClimate = COL_TC, `ERA5-Land` = COL_E5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  facet_wrap(~ source, ncol = 1) +
  labs(title    = "Drought episodes (scPDSI <= -1, length >= 3 months)",
       subtitle = "Bar height = consecutive months below -1; horizontal extent = episode span",
       x = NULL, y = "Episode length (months)") +
  theme(legend.position = "none")
g_episodes
```

**Reading.** Both products identify the same set of structural drought
episodes (timing, ordering, ranking by length). What differs is the
*count near the boundary*: short episodes that just dip below $-1$ in
TerraClimate may stay above $-1$ in ERA5-Land, inflating the count
discrepancy. This is the episode-level manifestation of the class-agreement
issue diagnosed in §4.2.

---

# 5. Formal stationarity testing

We apply the Kwiatkowski–Phillips–Schmidt–Shin (1992) protocol — two
*complementary* nulls, read jointly — exactly as in §4 of the SPI notebook.

| test | $H_0$ | rejection at 5 % means |
|---|---|---|
| ADF (`tseries::adf.test`)   | unit root (non-stationary)        | evidence of stationarity |
| KPSS (`tseries::kpss.test`) | level- (or trend-) stationarity   | evidence of non-stationarity |

```{r stationarity}
y_tc <- scpdsi$scPDSI_TerraClimate
y_e5 <- scpdsi$scPDSI_ERA5

run_stat <- function(y, name) {
  adf_res  <- tseries::adf.test(as.numeric(y), alternative = "stationary")
  kpss_l   <- tseries::kpss.test(as.numeric(y), null = "Level")
  kpss_t   <- tseries::kpss.test(as.numeric(y), null = "Trend")
  tibble(
    series    = name,
    test      = c("ADF",        "KPSS-level",   "KPSS-trend"),
    null      = c("unit root",  "level-stat.",  "trend-stat."),
    statistic = c(as.numeric(adf_res$statistic),
                  as.numeric(kpss_l$statistic),
                  as.numeric(kpss_t$statistic)),
    p_value   = c(adf_res$p.value, kpss_l$p.value, kpss_t$p.value),
    decision_5 = c(if (adf_res$p.value  < 0.05) "reject H0" else "do not reject H0",
                   if (kpss_l$p.value   < 0.05) "reject H0" else "do not reject H0",
                   if (kpss_t$p.value   < 0.05) "reject H0" else "do not reject H0")
  )
}

stat_tbl <- bind_rows(
  run_stat(y_tc, "TerraClimate"),
  run_stat(y_e5, "ERA5-Land")
)
knitr::kable(stat_tbl, digits = 3,
             caption = "ADF and KPSS tests on both scPDSI series (joint reading).")
```

**Joint reading and structural argument.**

* The canonical *stationarity verdict* is "reject ADF, do not reject KPSS":
  both tests agree.
* The canonical *unit-root verdict* is "do not reject ADF, reject KPSS".
* Mixed outcomes ("reject both" / "reject neither") are *inconclusive* by
  the test pair alone and is best resolved by a structural argument.

The structural argument here is decisive: **scPDSI is bounded by
construction**. The Palmer recursion saturates the index at approximately
$\pm 4$ via the duration factors, and the self-calibration step keeps the
extreme tails from accumulating. A genuine $I(1)$ process cannot be bounded;
therefore an observed ADF non-rejection on this series is more parsimoniously
interpreted as a *power problem* (the strong AR(1) component below leaves
little energy for the test to identify a unit root) than as evidence of
integration.

The modeling implication is that scPDSI should be modelled in **levels**, not
in differences. Forced first-differencing would amount to over-differencing
a strongly AR(1) process — a classic mis-specification that would whiten the
series and remove precisely the persistence structure we want to model.

---

# 6. Time-series structure — quantitative ACF/PACF

```{r ts_obj}
y_ts <- ts(y_tc,
           start = c(year(SPI_START), month(SPI_START)),
           frequency = 12)

# Sample autocorrelations at the first 12 lags, for quantitative reading.
acf_vals  <- as.numeric(acf(y_tc,  lag.max = 12, plot = FALSE)$acf[-1])
pacf_vals <- as.numeric(pacf(y_tc, lag.max = 12, plot = FALSE)$acf)
acf_tbl <- tibble(
  lag = 1:12,
  ACF  = acf_vals,
  PACF = pacf_vals
)
knitr::kable(acf_tbl, digits = 3,
             caption = "Sample ACF and PACF of scPDSI (TerraClimate) at lags 1-12.")
```

```{r acf_pacf_fig, fig.height = 3.8}
g_acf  <- ggAcf(y_ts,  lag.max = 36) +
  ggtitle("ACF — scPDSI (TerraClimate)") +
  THESIS_THEME
g_pacf <- ggPacf(y_ts, lag.max = 36) +
  ggtitle("PACF — scPDSI (TerraClimate)") +
  THESIS_THEME
g_acf | g_pacf
```

```{r roll_moments, fig.height = 3.0}
roll <- scpdsi %>%
  transmute(date,
            roll_mean = zoo::rollmean(y_tc, k = 12, fill = NA, align = "right"),
            roll_sd   = zoo::rollapply(y_tc, width = 12, FUN = sd,
                                       fill = NA, align = "right"))
g_roll <- ggplot(roll, aes(date)) +
  geom_line(aes(y = roll_mean, colour = "12-month mean"), linewidth = 0.5) +
  geom_line(aes(y = roll_sd,   colour = "12-month sd"),   linewidth = 0.5) +
  scale_colour_manual(values = c("12-month mean" = COL_TC,
                                  "12-month sd"   = COL_E5)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title    = "Rolling first and second moments (window = 12)",
       subtitle = "Visual stationarity screen for scPDSI (TerraClimate)",
       x = NULL, y = NULL, colour = NULL)
g_roll
```

**Quantitative reading.**

* The lag-1 sample autocorrelation is
  $\hat{\rho}_1 = `r round(acf_vals[1], 3)`$. This places monthly scPDSI in
  the strongly persistent regime characteristic of basin water-balance
  indices (typical values $0.85$–$0.95$), in stark contrast with
  $\mathrm{SPI}_{6}$ whose lag-1 autocorrelation is in the
  $0$–$0.3$ range — see §3 of the SPI notebook.
* The ACF *decays slowly* but *monotonically*; it does not oscillate, which
  rules out a strong seasonal component at the monthly scale and is
  consistent with a near-AR(1) structure.
* The PACF *cuts off sharply after lag 1*
  ($\hat{\phi}_{11} = `r round(pacf_vals[1], 3)`$,
   $|\hat{\phi}_{kk}| \ll \hat{\phi}_{11}$ for $k \ge 2$), which is the
  textbook diagnostic for an AR(1) process. Higher-order PACF spikes at
  multiples of 12 would indicate residual seasonality — they are absent
  here.

> **Persistence statement.** *The scPDSI series exhibits strong persistence
> consistent with its recursive water balance formulation, implying a
> clear autoregressive structure that should be explicitly accounted for in
> time-series modeling.*

This is **not** an artefact of the index construction; it is the direct
mathematical consequence of the soil-bucket recursion, in which the current
month's moisture state is a deterministic function of the previous state
plus current forcings.

---

# 7. Residual diagnostics from a baseline AR(1) fit

A baseline AR(1) is fit to the primary series, **not** as a forecasting
candidate, but as a *diagnostic anchor*: if the residuals from this minimal
specification already behave like white noise, then the AR(1) representation
captures essentially all of the linear memory and Chapter 4's SARIMA grid
need only refine seasonal terms; if they do not, the residual structure tells
us *what kind of additional terms* are required.

```{r ar1_fit}
fit_ar1 <- forecast::Arima(y_ts, order = c(1, 0, 0), include.mean = TRUE)
ar1_phi    <- as.numeric(coef(fit_ar1)["ar1"])
ar1_mu     <- as.numeric(coef(fit_ar1)["intercept"])
ar1_sigma  <- sqrt(fit_ar1$sigma2)

ar1_summary <- tibble(
  parameter = c("phi (AR1)", "mu (intercept)", "sigma (innovation sd)",
                "log-likelihood", "AIC", "BIC"),
  value     = c(ar1_phi, ar1_mu, ar1_sigma,
                as.numeric(logLik(fit_ar1)), AIC(fit_ar1), BIC(fit_ar1))
)
knitr::kable(ar1_summary, digits = 3,
             caption = "Baseline AR(1) fit on scPDSI (TerraClimate).")
```

```{r ar1_residuals, fig.height = 4.5}
res <- residuals(fit_ar1)

g_res_ts  <- autoplot(res) +
  geom_hline(yintercept = 0, linetype = 2, colour = COL_GREY) +
  ggtitle("AR(1) residuals over time") +
  ylab("residual") + xlab(NULL) + THESIS_THEME

g_res_acf <- ggAcf(res, lag.max = 36) +
  ggtitle("ACF of AR(1) residuals") + THESIS_THEME

g_res_qq <- ggplot(tibble(r = as.numeric(res)), aes(sample = r)) +
  stat_qq(size = 0.8, colour = COL_TC) +
  stat_qq_line(colour = COL_GREY, linetype = 2) +
  ggtitle("Normal Q-Q of AR(1) residuals") +
  xlab("Theoretical quantile") + ylab("Sample quantile") +
  THESIS_THEME

(g_res_ts | g_res_acf) / g_res_qq +
  plot_layout(heights = c(1, 1))
```

```{r ar1_tests}
# Ljung-Box test for residual autocorrelation at three diagnostic horizons
# (12, 24, 36 months). df is adjusted for the one estimated AR coefficient.
lb_tests <- bind_rows(lapply(c(12, 24, 36), function(L) {
  lb <- Box.test(res, lag = L, type = "Ljung-Box", fitdf = 1)
  tibble(test = "Ljung-Box", lag = L,
         statistic  = as.numeric(lb$statistic),
         p_value    = lb$p.value,
         decision_5 = if (lb$p.value < 0.05) "reject H0 (residual autocorr.)"
                      else "do not reject H0 (white-noise compatible)")
}))

# Jarque-Bera normality test on the residuals.
jb <- tseries::jarque.bera.test(as.numeric(res))
jb_tbl <- tibble(test = "Jarque-Bera",
                 statistic  = as.numeric(jb$statistic),
                 p_value    = jb$p.value,
                 decision_5 = if (jb$p.value < 0.05) "reject normality"
                              else "do not reject normality")

knitr::kable(lb_tests, digits = 3,
             caption = "Ljung-Box residual autocorrelation tests (fitdf = 1).")
knitr::kable(jb_tbl, digits = 3,
             caption = "Jarque-Bera test on AR(1) residuals.")
```

**Interpretation of the residual diagnostics.**

* The fitted lag-1 coefficient $\hat{\phi} = `r round(ar1_phi, 3)`$ is highly
  significant and consistent with the persistent regime diagnosed in §6.
* If the Ljung-Box statistics fail to reject the white-noise null at all
  three lags, the AR(1) specification is *adequate*: the linear memory is
  fully captured by a single autoregressive term and Chapter 4 can centre
  its SARIMA grid on $(1, 0, 0)$ with light seasonal extensions.
* If Ljung-Box rejects at any lag, the residuals retain structure and an
  MA term, a seasonal component, or both, would need to be admitted into the
  candidate set. The ACF panel above identifies the offending lag(s)
  visually.
* Jarque-Bera tests Gaussian innovations. Rejection does *not* invalidate
  the AR(1) point estimate (least-squares estimators of AR coefficients are
  consistent under non-Gaussianity) but it does tighten the case for
  bootstrap-based predictive intervals over Gaussian-asymptotic intervals,
  the route already taken in the SPI notebook.

---

# 8. AWC parameter-uncertainty propagation

scPDSI is non-linearly sensitive to the basin Available Water Capacity. The
sweep $\{90, 110, 120, 130, 150\}$ mm — explicitly $\pm 25\%$ around the
SoilGrids basin-mean of $120$ mm — was computed during the R driver step and
is propagated here as an explicit *parameter-uncertainty band*.

## 8.1 Quantification (not just visualisation)

```{r awc_load}
# §8.1 measures *pure* AWC uncertainty (not joint AWC × PET). The CSV now
# carries both PET methods (5 AWC × 2 PET × 312 months = 3120 rows) since
# the compute step was extended for §8.3's variance decomposition; restrict
# `sens` to TerraClimate here so the original AWC-only semantics of §8.1
# and §8.2 are preserved. The joint ensemble is loaded separately in §8.3.
sens <- readr::read_csv(SCPDSI_AWC_SWEEP, show_col_types = FALSE) %>%
  dplyr::filter(pet_method == "TerraClimate") %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

awc_summary <- sens %>%
  group_by(AWC_mm) %>%
  summarise(mean_scPDSI = mean(scPDSI, na.rm = TRUE),
            sd_scPDSI   = sd(scPDSI,   na.rm = TRUE),
            min_scPDSI  = min(scPDSI,  na.rm = TRUE),
            max_scPDSI  = max(scPDSI,  na.rm = TRUE),
            .groups = "drop")
knitr::kable(awc_summary, digits = 3,
             caption = "scPDSI moments under the AWC sweep (TerraClimate forcing).")

# Per-month spread across AWC values: how much does the AWC parameter
# perturb the index at each timestamp?
spread <- sens %>%
  group_by(date) %>%
  summarise(low   = min(scPDSI, na.rm = TRUE),
            high  = max(scPDSI, na.rm = TRUE),
            range = high - low,
            sd    = sd(scPDSI, na.rm = TRUE),
            .groups = "drop")

awc_uncertainty <- tibble(
  metric = c("Mean cross-AWC range",
             "Median cross-AWC range",
             "95th-percentile cross-AWC range",
             "Mean cross-AWC sd",
             "Range during deepest drought month",
             "Range during nearest-normal month"),
  value = c(mean(spread$range,   na.rm = TRUE),
            median(spread$range, na.rm = TRUE),
            quantile(spread$range, 0.95, na.rm = TRUE),
            mean(spread$sd, na.rm = TRUE),
            spread$range[which.min(spread$low)],
            spread$range[which.min(abs(scpdsi$scPDSI_TerraClimate))])
)
knitr::kable(awc_uncertainty, digits = 3,
             caption = "Quantified AWC parameter-uncertainty propagation.")
```

## 8.2 Envelope figure

```{r awc_envelope, fig.height = 4.5}
env <- spread %>%
  left_join(scpdsi %>% select(date, central = scPDSI_TerraClimate), by = "date")

g_envelope <- ggplot(sens, aes(date, scPDSI,
                                colour = factor(AWC_mm),
                                group  = AWC_mm)) +
  geom_line(linewidth = 0.4, alpha = 0.85) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1, name = "AWC (mm)") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title    = "TerraClimate scPDSI under the AWC sweep",
       subtitle = "AWC at 90 / 110 / 120 / 130 / 150 mm (+/-25% around the SoilGrids central value)",
       x = NULL, y = "scPDSI")

g_band <- ggplot(env, aes(date)) +
  geom_ribbon(aes(ymin = low, ymax = high), fill = COL_TC, alpha = 0.20) +
  geom_line(aes(y = central), colour = COL_TC, linewidth = 0.5) +
  geom_hline(yintercept = 0, linetype = 2, colour = COL_GREY) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title    = "Parameter-uncertainty band from the AWC sweep",
       subtitle = "Shaded = min/max across AWC; line = central trajectory (AWC = 120 mm)",
       x = NULL, y = "scPDSI")

g_envelope / g_band
```

**Interpretation.**

* The mean cross-AWC range of $`r round(mean(spread$range), 3)`$ scPDSI units
  sets an *empirical floor* on differences between scPDSI variants: any
  model-class comparison in Chapter 4 producing forecasts within this band
  cannot be considered significantly different on AWC-uncertainty grounds
  alone.
* The envelope is **tightest at the saturation extremes and widest in the
  dynamic mid-range**: range during the deepest drought month is
  $`r round(spread$range[which.min(spread$low)], 3)`$ scPDSI units, while
  range during the nearest-normal month is
  $`r round(spread$range[which.min(abs(scpdsi$scPDSI_TerraClimate))], 3)`$.
  This reverses the intuitive expectation that AWC matters most during
  droughts, and it has a clean physical reading: at deep-drought
  saturation the soil-bucket is empty under every AWC variant, so the
  index is anchored by the Palmer ceiling rather than by AWC; in
  near-normal months the bucket still has room to deplete or recharge, so
  AWC determines how quickly that happens and the variants diverge.
* The PET-method choice is the main axis of inter-variant uncertainty,
  not AWC: the TerraClimate-vs-ERA5 RMSE in §4
  ($`r round(rmse_idx, 3)`$ scPDSI units) is roughly $3 \times$ the mean
  cross-AWC range. Practically, AWC uncertainty should still be *propagated*
  into Chapter 4's predictive intervals, but it should not be confused with
  the larger PET-method uncertainty band reported in §4.2.

## 8.3 Variance decomposition: AWC vs PET attribution

The §8.1 envelope quantified AWC uncertainty in isolation. Under the
post-pivot architecture (Chapter 1.4.2), scPDSI uncertainty has *two*
meaningful axes — AWC parameter and PET-method choice — because the
shared-precipitation backbone makes precipitation contribute zero variance
to inter-variant disagreement by construction. The decomposition below uses
the joint AWC × PET ensemble (5 AWC × 2 PET × 312 months = 3120 scPDSI
values from `scpdsi_awc_sensitivity.csv`) to partition the per-month
cross-variant variance into AWC, PET, and their interaction, then averages
across the record.

```{r variance_decomposition}
# ---- 8.3.1  Load joint AWC x PET ensemble ----
sens_full <- readr::read_csv(SCPDSI_AWC_SWEEP, show_col_types = FALSE) %>%
  dplyr::mutate(date = as.Date(date))

stopifnot(
  nrow(sens_full) == 312L * 5L * 2L,
  setequal(unique(sens_full$AWC_mm),     c(90, 110, 120, 130, 150)),
  setequal(unique(sens_full$pet_method), c("TerraClimate", "ERA5-Land"))
)

# ---- 8.3.2  Per-month two-way ANOVA without replication ----
# At each date t, 10 scPDSI values arranged as a 5 (AWC) x 2 (PET) grid.
#   SS_total = sum over 10 cells of (cell - grand_mean)^2
#   SS_AWC   = (#PET) * sum over AWC marginals of (mean_a - grand_mean)^2
#   SS_PET   = (#AWC) * sum over PET marginals of (mean_p - grand_mean)^2
#   SS_int   = SS_total - SS_AWC - SS_PET   (residual / interaction)
n_pet <- length(unique(sens_full$pet_method))
n_awc <- length(unique(sens_full$AWC_mm))

decomp_per_month <- sens_full %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(
    grand    = mean(scPDSI),
    SS_total = sum((scPDSI - grand)^2),
    SS_AWC   = {
      m <- tapply(scPDSI, AWC_mm,     mean)
      n_pet * sum((m - grand)^2)
    },
    SS_PET   = {
      m <- tapply(scPDSI, pet_method, mean)
      n_awc * sum((m - grand)^2)
    },
    .groups  = "drop"
  ) %>%
  dplyr::mutate(SS_int = pmax(SS_total - SS_AWC - SS_PET, 0))

# ---- 8.3.3  Aggregate variance attribution over the record ----
totals <- decomp_per_month %>%
  dplyr::summarise(dplyr::across(dplyr::starts_with("SS_"), sum))

variance_attribution <- tibble::tibble(
  Source       = c("PET method (Penman--Monteith vs Thornthwaite)",
                   "AWC parameter (5 values: 90, 110, 120, 130, 150 mm)",
                   "Interaction (AWC \times PET)"),
  `Total SS`   = c(totals$SS_PET, totals$SS_AWC, totals$SS_int),
  `% of total` = round(100 * c(totals$SS_PET, totals$SS_AWC, totals$SS_int)
                       / totals$SS_total, 1)
)

knitr::kable(
  variance_attribution,
  digits  = c(0, 2, 1),
  caption = paste(
    "Variance decomposition of scPDSI across the joint AWC x PET ensemble.",
    "Per-month sums of squares summed over the 312-month record.",
    "Precipitation contributes zero by construction (shared station backbone)."
  )
)

readr::write_csv(variance_attribution,
                 file.path(OUT_DIR, "variance_decomposition_awc_pet.csv"))
```

**Reading the decomposition.** PET-method choice (Penman--Monteith vs
Thornthwaite) absorbs the largest share of cross-variant scPDSI variance
($`r round(100 * sum(decomp_per_month$SS_PET) / sum(decomp_per_month$SS_total), 1)`\%$
of total SS). AWC parameter variation contributes
$`r round(100 * sum(decomp_per_month$SS_AWC) / sum(decomp_per_month$SS_total), 1)`\%$,
and a non-trivial
$`r round(100 * sum(decomp_per_month$SS_int) / sum(decomp_per_month$SS_total), 1)`\%$
is captured by the AWC $\times$ PET interaction term. The interaction
magnitude is the methodologically interesting finding: it is *larger* than
the AWC main effect, signalling that AWC sensitivity is not separable from
PET method. Per-cell summary statistics confirm the asymmetry — the
AWC-driven spread in ERA5-Land cell means is approximately $0.37$ scPDSI
units (AWC $= 90$: $-0.25$, AWC $= 150$: $+0.12$), while in TerraClimate
cells the same spread is only $0.10$ units (AWC $= 90$: $+0.12$, AWC $=
150$: $+0.09$). The physical interpretation is direct: under Thornthwaite's
lower atmospheric demand, the soil bucket spends more time in the
partially-filled regime where its capacity (AWC) actively governs the
water-balance state; under Penman--Monteith's higher demand the bucket
saturates near empty more frequently, partially neutralising AWC's effect
on the index. The §4 decision to retain TerraClimate (Penman--Monteith)
as primary therefore stands; ERA5-Land (Thornthwaite) remains a robustness
check, with the additional caveat that the AWC sensitivity reported in
§8.1 should be read as conditional on the TerraClimate forcing chain
(taken as the reference PET method), and the interaction term documented
here is what justifies *not* assuming AWC uncertainty transports
unchanged across PET methods.

## 8.4 Composite uncertainty band: AWC nested inside PET

The composite figure below renders the central scPDSI series (TerraClimate,
AWC = 120 mm) with two nested uncertainty bands derived from the joint
ensemble: the **wider PET-method band** (TC vs ERA5 envelope at central
AWC) wrapping the **narrower AWC band** (5-AWC range at TC central PET).
Reading the visual — any forecast disagreement falling within the inner
band is AWC-uncertainty noise; any falling within the outer band but
outside the inner is PET-method noise; any falling outside the outer band
is interpretable signal.

```{r composite_uncertainty_band, fig.height = 5}
# AWC band (TC PET, all 5 AWC values, per month min and max)
awc_band_tc <- sens_full %>%
  dplyr::filter(pet_method == "TerraClimate") %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(awc_lo = min(scPDSI),
                   awc_hi = max(scPDSI),
                   .groups = "drop")

# PET-method band (central AWC = 120, both PET methods, from scpdsi_master)
pet_band <- scpdsi %>%
  dplyr::transmute(date,
                   pet_lo = pmin(scPDSI_TerraClimate, scPDSI_ERA5),
                   pet_hi = pmax(scPDSI_TerraClimate, scPDSI_ERA5))

# Central line: primary = TerraClimate at central AWC
central_line <- scpdsi %>% dplyr::select(date, value = scPDSI_TerraClimate)

g_composite <- ggplot() +
  geom_ribbon(data = pet_band,
              aes(date, ymin = pet_lo, ymax = pet_hi,
                  fill = "PET-method band (TC vs ERA5, AWC = 120)"),
              alpha = 0.28) +
  geom_ribbon(data = awc_band_tc,
              aes(date, ymin = awc_lo, ymax = awc_hi,
                  fill = "AWC band (TC PET, AWC in {90, 110, 120, 130, 150})"),
              alpha = 0.55) +
  geom_line(data = central_line,
            aes(date, value, colour = "Primary scPDSI (TC, AWC = 120)"),
            linewidth = 0.5) +
  geom_hline(yintercept = c(-2, 2), linetype = "dotted", colour = "grey50") +
  geom_hline(yintercept = 0,        linetype = "dashed", colour = "grey60") +
  scale_fill_manual(name = NULL,
                    values = c("PET-method band (TC vs ERA5, AWC = 120)" = COL_E5,
                               "AWC band (TC PET, AWC in {90, 110, 120, 130, 150})" = COL_TC)) +
  scale_colour_manual(name = NULL,
                      values = c("Primary scPDSI (TC, AWC = 120)" = "black")) +
  labs(title    = "scPDSI composite uncertainty band",
       subtitle = paste0("PET-method band (wider, red) wraps AWC band (narrower, blue); ",
                          "central line = primary"),
       x = NULL, y = "scPDSI") +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE),
         colour = guide_legend(nrow = 1))
g_composite

ggsave(file.path(FIG_DIR, "fig_composite_uncertainty_band.pdf"),
       g_composite, width = 11, height = 5)
```

**Reading the composite figure.** Three visual properties are interpretable.
First, the AWC band (blue) is consistently narrower than the PET band (red):
this is the variance-decomposition result of §8.3 expressed visually.
Second, both bands tighten toward the saturation extremes (deep droughts
near the Palmer floor, wet phases at the ceiling) and widen in the dynamic
mid-range — the same envelope-reversal pattern §8.1 quantified for AWC
alone, here generalised across the joint ensemble. Third, the central
TerraClimate line lies inside the PET band by construction but is
*systematically positioned at the lower (drier) edge* during the
2000--2006 sustained drought (§10), reflecting Penman--Monteith's stronger
PET demand and consistent with the amplitude-compression result reported
in §4.2 (OLS slope $0.834$, ERA5 amplitude $\approx 83\%$ of TC).


---

# 9. Visualisation integration

The pre-computed PDF figures stored in `03_outputs/M3_scpdsi_diagnostics/figures/`
are the canonical thesis figures and are referenced rather than re-rendered:

* `fig_scpdsi_timeseries.pdf` — reproduced as `g_ts` (§3.2). Reads as: "do
  the two indices follow the same drought trajectory?"
* `fig_validation_tc_vs_era5.pdf` — reproduced as `g_scatter + g_resid`
  (§4.3). Reads as: "where does the disagreement live — slope, offset, or
  noise?"
* `fig_scpdsi_drought_episodes.pdf` — reproduced as `g_episodes` (§4.4).
  Reads as: "are the same episodes detected, at compatible durations?"
* `fig_scpdsi_awc_sensitivity.pdf` — reproduced as `g_envelope / g_band`
  (§8.2). Reads as: "how much of the index is the index, and how much is the
  AWC parameter?"

```{r list_figs, results = "asis"}
figs <- list.files(FIG_DIR, pattern = "\\.pdf$", full.names = TRUE)
if (length(figs)) {
  cat("**Canonical figures on disk:**\n\n")
  for (f in figs) cat("- `", basename(f), "`\n", sep = "")
} else {
  cat("_No PDF figures found in_ ", FIG_DIR, "\n")
}
```

---

# 10. Historical drought validation against documented episodes

The four major drought episodes documented in the Algerian hydroclimate
literature for the 1990–2015 study window are matched here against the
SPI-6 and scPDSI signatures detected in the data, using the same run-length
threshold convention applied in §4.4 (index $\le -1$ for at least 3
consecutive months). Each documented episode is either reproduced by the
indices or not, and the comparative signatures across the two indices are
interpretable in their own right.

```{r drought_event_validation}
# ---- 10.1  Documented Chélif/Algerian drought episodes ----
docs <- tibble(
  Episode      = c("1992–1993", "1999", "2000–2006", "2014–2015"),
  window_start = as.Date(c("1992-01-01","1999-01-01","2000-01-01","2014-01-01")),
  window_end   = as.Date(c("1993-12-01","1999-12-01","2006-12-01","2015-12-01")),
  Citation     = c(
    "Meddi & Hubert (2003); Meddi et al. (2010)",
    "Meddi et al. (2010); Zeroual et al. (2016)",
    "Meddi et al. (2010); Zeroual et al. (2016); Cook et al. (2018)",
    "Cook et al. (2018); Zeroual et al. (2016)"
  )
)

# ---- 10.2  Episode detector with peak (extends episodes_of() of §4.4) ----
detect_with_peak <- function(x, dates, threshold = -1.0, min_length = 3L) {
  flag <- !is.na(x) & x <= threshold
  if (!any(flag))
    return(tibble(start = as.Date(character()), end = as.Date(character()),
                  length_months = integer(), peak = numeric()))
  rle_obj <- rle(flag)
  ends    <- cumsum(rle_obj$lengths)
  starts  <- ends - rle_obj$lengths + 1L
  out <- tibble(
    start         = dates[starts][rle_obj$values],
    end           = dates[ends][rle_obj$values],
    length_months = as.integer(rle_obj$lengths[rle_obj$values])
  ) %>% dplyr::filter(length_months >= min_length)
  out$peak <- vapply(seq_len(nrow(out)), function(k) {
    seg <- x[which(dates >= out$start[k] & dates <= out$end[k])]
    round(min(seg, na.rm = TRUE), 2)
  }, numeric(1))
  out
}

ep_spi6 <- detect_with_peak(spi_raw$SPI_6_regional,     spi_raw$date)
ep_tc   <- detect_with_peak(scpdsi$scPDSI_TerraClimate, scpdsi$date)
ep_e5   <- detect_with_peak(scpdsi$scPDSI_ERA5,         scpdsi$date)

# ---- 10.3  Match each documented episode to overlapping data episodes ----
# When a documented window contains multiple short events (e.g. SPI-6 in
# the 1992–1993 window), we report: earliest start, total months, deepest peak.
match_in_window <- function(eps, w0, w1) {
  hit <- dplyr::filter(eps, end >= w0, start <= w1)
  if (nrow(hit) == 0)
    return(list(start = as.Date(NA), n_mo = NA_integer_, peak = NA_real_))
  list(start = min(hit$start),
       n_mo  = sum(hit$length_months),
       peak  = min(hit$peak))
}

fmt_date <- function(d) ifelse(is.na(d), "—", format(d, "%Y-%m"))
fmt_dur  <- function(n) ifelse(is.na(n), "—", paste0(n, " mo"))
fmt_val  <- function(v) ifelse(is.na(v), "—", sprintf("%.2f", v))

table_rows <- purrr::map_dfr(seq_len(nrow(docs)), function(i) {
  s <- match_in_window(ep_spi6, docs$window_start[i], docs$window_end[i])
  t <- match_in_window(ep_tc,   docs$window_start[i], docs$window_end[i])
  e <- match_in_window(ep_e5,   docs$window_start[i], docs$window_end[i])
  tibble(
    Episode                = docs$Episode[i],
    `SPI-6 onset`          = fmt_date(s$start),
    `SPI-6 duration`       = fmt_dur(s$n_mo),
    `SPI-6 peak`           = fmt_val(s$peak),
    `scPDSI onset (TC)`    = fmt_date(t$start),
    `scPDSI duration (TC)` = fmt_dur(t$n_mo),
    `scPDSI peak (TC)`     = fmt_val(t$peak),
    `scPDSI peak (ERA5)`   = fmt_val(e$peak),
    Citation               = docs$Citation[i]
  )
})

knitr::kable(
  table_rows,
  caption = paste(
    "Historical drought validation: documented Chélif/Algerian episodes",
    "(literature anchor) matched against detected episodes in SPI-6 and scPDSI.",
    "Onset = first month meeting the run-length criterion (index <= -1 for >= 3",
    "consecutive months); duration = total months in the matched episode(s);",
    "peak = lowest index value reached during the episode.",
    "Em-dash marks an episode the index does not detect under the threshold."
  )
)

# Persist for cross-referencing across chapters
readr::write_csv(table_rows, file.path(OUT_DIR, "drought_event_validation.csv"))
```

**Reading the table.** All four documented episodes are reproduced by at
least one index in the data, with no false negatives. The relative SPI-6
vs scPDSI signatures across the four episodes are coherent with the
methodological characterisation established in §4 and reveal the memory
asymmetry that the two-index framework was designed to expose.

**1992–1993 — early-decade major drought.** SPI-6 records the period as
13 cumulative months of below-threshold conditions (peak $-2.00$);
scPDSI records a single 20-month sustained event reaching the Palmer
floor at $-4.11$ (TerraClimate) and $-4.06$ (ERA5-Land). This episode
forms part of the long-term Northwest Algerian precipitation decline
documented by Meddi & Hubert (2003), and is the first instance in the
record where the soil-bucket recursion accumulates a precipitation
deficit an order of magnitude more strongly than the standardised-anomaly
framework smooths it.

**1999 — short, intense meteorological drought.** SPI-6 reaches its
record minimum of $-3.33$ in June 1999 — the deepest SPI signal in the
entire 1990–2015 series — across an 8-month event driven by an
exceptionally dry winter–spring 1999. Both scPDSI variants identify the
period as an 11-month event with peaks of $-2.96$ (TC) and $-3.42$ (ERA5).
The fact that SPI-6 records a *deeper* peak than scPDSI for this single
event is methodologically informative: the meteorological deficit is so
abrupt that the water-balance recursion has not yet had time to
accumulate to the Palmer floor, while the standardised-anomaly framework
reports the precipitation extremity directly.

**2000–2006 — sustained dry decade (the chapter's headline finding).**
SPI-6 detects this six-year period as multiple discrete short episodes
(the longest a 6-month event in 2001 with peak $-1.70$), reflecting that
individual seasons crossed back above the $-1$ threshold even as the
basin remained in cumulative deficit. scPDSI captures it as a *single
continuous episode*: 72 months under TerraClimate (peak $-4.01$ in June
2001) and 48 months under ERA5-Land (peak $-4.08$ in October 2001).
This six-year contrast is the single most physically interpretable
finding in the chapter — the same hydroclimatic record diagnosed as
"intermittent drought" by a memoryless statistical index and as
"sustained drought" by a recursive water-balance index, with neither
diagnosis wrong but each answering a different question. Documented
across Meddi et al. (2010), Zeroual et al. (2016), and framed within
the regional Mediterranean intensification of Cook et al. (2018).

**2014–2015 — return-to-deficit episode at end of record.** After the
2008–2013 wet recovery, the record closes with a return to negative
conditions. SPI-6 detects a brief 3-month event (peak $-1.54$); scPDSI
ERA5-Land captures a 15-month sustained episode (peak $-2.62$); scPDSI
TerraClimate registers a shorter 3-month signal (peak $-2.07$). The
asymmetry between the two scPDSI variants is informative in its own
right: ERA5-Land's lower Thornthwaite PET demand combined with the long
memory of the recursion produces a sustained negative state, while
TerraClimate's higher Penman–Monteith demand permits the index to
recover more readily once precipitation rebounds in late 2015. This
episode is treated as a leading indicator of the post-2010 Mediterranean
climate shift discussed in Cook et al. (2018) and Zeroual et al. (2016).

## 10.4 Lag-correlation between SPI-6 and scPDSI: temporal coupling

The validation against documented episodes (§10.1–§10.3) characterises
*which* events the two indices co-detect; the cross-correlation
analysis below characterises *when* they co-detect them. We compute
$\rho_{xy}(k) = \mathrm{Cor}(\mathrm{SPI}_{6,t}, \mathrm{scPDSI}_{t+k})$
at lags $k \in [-12, +12]$ months. Following R's `ccf(x, y)` convention
$\rho_{xy}(k) = \mathrm{Cor}(x_{t+k}, y_t)$, **negative $k$ means SPI-6
leads scPDSI** by $|k|$ months (older SPI predicts current scPDSI);
**positive $k$ means scPDSI leads SPI-6** by $k$ months (older scPDSI
predicts current SPI).

```{r lag_correlation, fig.height = 4.5}
# ---- 10.4.1  Align series on the common date axis ----
ccf_data <- spi_raw %>%
  dplyr::select(date, SPI_6_regional) %>%
  dplyr::inner_join(scpdsi %>% dplyr::select(date,
                                              scPDSI_TerraClimate,
                                              scPDSI_ERA5),
                    by = "date") %>%
  tidyr::drop_na()

n_eff <- nrow(ccf_data)
ci_95 <- 1.96 / sqrt(n_eff)

# ---- 10.4.2  Cross-correlation at lags -12..+12 ----
ccf_tc <- stats::ccf(ccf_data$SPI_6_regional, ccf_data$scPDSI_TerraClimate,
                     lag.max = 12, plot = FALSE)
ccf_e5 <- stats::ccf(ccf_data$SPI_6_regional, ccf_data$scPDSI_ERA5,
                     lag.max = 12, plot = FALSE)

ccf_df <- dplyr::bind_rows(
  tibble(lag = as.integer(ccf_tc$lag),
         r   = as.numeric(ccf_tc$acf),
         source = "TerraClimate"),
  tibble(lag = as.integer(ccf_e5$lag),
         r   = as.numeric(ccf_e5$acf),
         source = "ERA5-Land")
)

# Peak per source
peak_df <- ccf_df %>%
  dplyr::group_by(source) %>%
  dplyr::slice_max(r, n = 1) %>%
  dplyr::ungroup()

# ---- 10.4.3  Render the panel ----
g_ccf <- ggplot(ccf_df, aes(lag, r, colour = source)) +
  annotate("rect", xmin = -Inf, xmax = Inf,
           ymin = -ci_95, ymax = ci_95,
           alpha = 0.15, fill = "grey60") +
  geom_hline(yintercept = 0, colour = COL_GREY, linewidth = 0.4) +
  geom_segment(aes(xend = lag, yend = 0),
               linewidth = 0.6, alpha = 0.55,
               position = position_dodge(width = 0.45)) +
  geom_point(size = 2.4,
             position = position_dodge(width = 0.45)) +
  geom_point(data = peak_df, aes(lag, r),
             shape = 21, size = 4.5, stroke = 1.0, fill = NA,
             position = position_dodge(width = 0.45)) +
  scale_colour_manual(values = c("TerraClimate" = COL_TC,
                                  "ERA5-Land"    = COL_E5)) +
  scale_x_continuous(breaks = seq(-12, 12, by = 2)) +
  labs(title    = "Cross-correlation of SPI-6 and scPDSI",
       subtitle = sprintf(paste0(
         "Grey band = 95%% CI (±%.3f, n = %d). ",
         "Circled point = peak. k > 0: SPI leads scPDSI; k < 0: scPDSI leads SPI."),
         ci_95, n_eff),
       x = "Lag k (months)",
       y = expression(rho[xy](k)),
       colour = NULL)
g_ccf

# Persist diagnostic CSV for cross-chapter reference
readr::write_csv(ccf_df, file.path(OUT_DIR, "lag_correlation_spi6_scpdsi.csv"))
```

**Reading the lag-correlation panel.** The cross-correlation function
exposes three properties of the SPI-6/scPDSI temporal coupling that the
concurrent validation of §4 cannot capture.

*Concurrent peak.* The maximum cross-correlation occurs at $k = 0$:
$\rho_{xy}(0) = 0.63$ (TerraClimate) and $0.66$ (ERA5-Land), both well
above the 95 % significance band (±$0.11$). The two indices co-vary
most strongly in the *same* month, consistent with both being derived
from the same precipitation backbone integrated over overlapping
look-back windows.

*Symmetric immediate neighbourhood.* For $|k| \le 3$, the
cross-correlation function is approximately symmetric:
$\rho_{xy}(\pm 1) \approx 0.58$, $\rho_{xy}(\pm 2) \approx 0.52$,
$\rho_{xy}(\pm 3) \approx 0.45$. Within a quarterly horizon, neither
index leads the other in any operationally meaningful sense.

*Asymmetric long tail.* For $|k| \ge 6$, the function becomes markedly
asymmetric. At $k = -12$ (SPI-6 leads scPDSI by 12 months),
$\rho_{xy} = 0.28$ (TC) and $0.35$ (ERA5); at $k = +12$ (scPDSI leads
SPI-6 by 12 months), $\rho_{xy} = 0.17$ (TC) and $0.17$ (ERA5),
near the 95 % band. A precipitation anomaly captured by SPI-6 still
influences scPDSI a year later, but a year-old scPDSI state is
near-uncorrelated with current SPI. This is the physical signature of
the two-index framework: scPDSI inherits SPI's memory through the
water-balance recursion, while SPI is reset each window and carries no
memory of the prior scPDSI state. The two indices are *causally
asymmetric* despite their concurrent covariation — a property that
justifies treating them as separate stochastic processes (Chapter 1,
§1.4.3) for the modelling phase of Chapter 4.


---

# 11. Transition to the modeling phase (Chapter 4)

This section converts the diagnostics above into modeling-side decisions.
The chain *diagnostic -> consequence* is laid out explicitly so that
Chapter 4 can be audited against it.

| Diagnostic (this notebook) | Modeling consequence |
|---|---|
| §5: ADF/KPSS verdict + bounded scPDSI by construction | Model in **levels**, not in differences; do not force $d \ge 1$. |
| §6: ACF slow decay; PACF cut-off at lag 1 ($\hat{\phi}_{11} \approx `r round(pacf_vals[1], 2)`$) | Centre the SARIMA grid on $(1, 0, 0)$; allow seasonal $(P, Q)_{12}$ extensions only. |
| §7: Ljung-Box on AR(1) residuals | If white, AR(1) is sufficient; if not, admit MA / seasonal terms guided by the offending lag. |
| §7: Jarque-Bera on AR(1) residuals | Use **bootstrap predictive intervals** rather than Gaussian-asymptotic intervals. |
| §8: Mean cross-AWC range = $`r round(mean(spread$range), 3)`$ | Add an **AWC-ensemble axis** to predictive uncertainty in Chapter 4. |
| §4: TC-vs-ERA5 amplitude bias | TerraClimate as **primary**; ERA5 as **robustness check**; report both. |
| §1.3: SPI / scPDSI separation policy | Do **not** merge the two series at the data layer; any joint analysis is performed inside the chapter that needs it, with an explicit hypothesis. |

**Why scPDSI requires autoregressive modeling.** The strong AR(1) signature
diagnosed in §6 is not a statistical artefact; it is the linear projection
of the recursive water balance underlying the Palmer formulation. A non-AR
specification (e.g. $\mathrm{ARIMA}(0, 0, 1)$ or pure seasonal) is
*structurally inconsistent* with the index's data-generating process. AR(1)
is the parsimonious lower bound; SARIMA $(p, 0, q) \times (P, 0, Q)_{12}$
with small $p, q, P, Q$ is the natural enrichment.

**Why scPDSI is modelled independently of SPI.** SPI is a memoryless
statistical transform; scPDSI is a recursive water-balance state. A naive
multivariate model (e.g., VAR with shared error covariance) on the joint
vector $(\mathrm{SPI}_{6, t}, \mathrm{scPDSI}_t)$ would absorb the scPDSI
persistence into the cross-covariance matrix and bias the SPI dynamics
towards spurious persistence. Independent univariate models preserve each
process's intrinsic memory; a joint analysis, when desired, is then
performed *between fitted predictives*, not between raw series.

The objects ready to flow into Chapter 4 are:

* `scpdsi$scPDSI_TerraClimate` — primary series (length 312).
* `scpdsi$scPDSI_ERA5` — validation series (length 312).
* `sens` — AWC sensitivity ensemble (5 trajectories x 312 months).
* `meta` — JSON provenance for the methodological appendix.

---

# 12. Conclusion

* **Validated.** Structural alignment with the SPI master index is
  exact; TerraClimate-vs-ERA5 dynamics agree at $r = 0.895$ with a
  $93\%$ hit rate on D2+ events under the post-pivot shared-precipitation
  architecture; the AWC envelope is bounded and interpretable; the four
  documented Chélif/Algerian drought episodes (§10) are reproduced by at
  least one index with no false negatives.
* **Uncertain.** Absolute amplitude differs between products (RMSE
  $1.009$, class agreement $0.47$), driven exclusively by the
  Penman-Monteith vs Thornthwaite PET contrast now that precipitation is
  shared. This uncertainty is reported, not resolved; it is the main
    source of provenance disagreement in Chapter 4.
* **Diagnosed.** scPDSI (TerraClimate) is strongly persistent
  ($\hat{\rho}_1 \approx `r round(acf_vals[1], 2)`$,
  $\hat{\phi}_{11} \approx `r round(pacf_vals[1], 2)`$), bounded by
  construction, and statistically compatible with stationarity in levels.
  Baseline AR(1) is the natural anchor; SARIMA grids are constrained to
  $(p, 0, q) \times (P, 0, Q)_{12}$ with small orders.
* **Ready for modeling.** Chapter 4 inherits a coherent, well-typed pair of
  drought series (one statistical, one physical), a parameter-uncertainty
  band quantified at $`r round(mean(spread$range), 3)`$ index units, and a
  transparent rule for how, and only how, the two may be combined.

```{r save_summary, include = FALSE}
# Persist the validation, AWC and AR(1) diagnostic tables so the report
# notebook can cross-reference them without re-running this script.
readr::write_csv(metrics_tbl,
                 file.path(OUT_DIR, "validation_metrics_chapter2b.csv"))
readr::write_csv(awc_summary,
                 file.path(OUT_DIR, "awc_sensitivity_summary.csv"))
readr::write_csv(awc_uncertainty,
                 file.path(OUT_DIR, "awc_uncertainty_quantified.csv"))
readr::write_csv(stat_tbl,
                 file.path(OUT_DIR, "stationarity_tests.csv"))
readr::write_csv(ar1_summary,
                 file.path(OUT_DIR, "ar1_baseline_summary.csv"))
readr::write_csv(lb_tests,
                 file.path(OUT_DIR, "ljung_box_residuals.csv"))
readr::write_csv(jb_tbl,
                 file.path(OUT_DIR, "jarque_bera_residuals.csv"))
```
